---
title: "Quality control"
author: "Katharina Hembach"
date: "5/26/2020"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE)
```

### Load packages
```{r, message = FALSE}
library(scater)
library(scales)
library(viridis)
```

### Load sce
```{r}
sce <- readRDS(file.path("output", "sce_01_preprocessing.rds"))
```

## Quality control
We compute cell-level QC.
```{r, message = FALSE}
(mito <- grep("MT-", rownames(sce), value = TRUE))
sce <- addPerCellQC(sce, subsets = list(Mt = mito))
# we compute the fraction of mitochondrial genes and the logit of it 
sce$subsets_Mt_fraction <- (sce$subsets_Mt_percent + 0.001) /100
sce$subsets_Mt_fraction_logit <- qlogis(sce$subsets_Mt_fraction + 0.001)
# library size
summary(sce$sum)
# number of detected genes per cell
summary(sce$detected)
# percentage of counts that come from mitochondrial genes:
summary(sce$subsets_Mt_percent)
```

### Diagnostic plots

The number of counts per cell:
```{r}
plotColData(sce, x = "sample_id", y = "sum")
```

The number of genes:
```{r}
plotColData(sce, x = "sample_id", y = "detected")
```

The percentage of mitochondrial genes:
```{r}
plotColData(sce, x = "sample_id", y = "subsets_Mt_percent")
```

We plot the total number of counts against the number of detected genes and color by the fraction of mitochondrial genes:
```{r}
cd <- data.frame(colData(sce))
ggplot(cd, aes(x = sum, y = detected, color = subsets_Mt_fraction)) +
  geom_point(alpha = 0.7) + 
  geom_density_2d(color = "grey", bins = 6) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~sample_id) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("sum of counts") + 
  ylab("number of detected genes") + 
  labs(color = "mitochondrial fraction") +
  scale_color_viridis(trans = "logit", breaks = c(0.01, 0.1, 0.25, 0.5, 0.75))
```

We plot the total number of counts against the mitochondrial content. Well-behaved cells should have many expressed genes and a low fraction of mitochondrial genes. High mitochondrial content indicates empty or damaged cells. 
```{r}
ggplot(cd, aes(x = sum, y = subsets_Mt_fraction)) +
  geom_point(color = "darkgrey", alpha = 0.3) + 
  geom_density_2d(color = "lightblue") +
  scale_x_log10() +
  scale_y_continuous(trans = 'logit', 
                     breaks = c(0.01, 0.05, 0.1, 0.2, 0.5, 0.75)) +
  facet_wrap(~sample_id) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("sum of counts") + 
  ylab("logit(mitochondrial fraction)")
```

We plot the top 20 genes with highest expression. Mitochondrial genes, actin, ribosomal proteins or MALAT1 are examples of genes that are expected to have very high expression. 
```{r, dev="png", eval = FALSE}
plotHighestExprs(sce, n = 20)
```
