---
title: "Filtering"
author: "Katharina Hembach"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE)
```

### Load packages
```{r, message = FALSE}
library(scater)
library(LSD)
library(dplyr)
library(edgeR)
library(ggrepel)
```

## Load data

```{r}
sce <- readRDS(file.path("output", "sce_02_quality_control.rds"))
```

## Identification of outlier cells

Based on the QC metrics, we now identify outlier cells:

```{r}
cols <- c("sum", "detected", "subsets_Mt_percent")
log <- c(TRUE, TRUE, FALSE)
type <- c("both", "both", "higher")

drop_cols <- paste0(cols, "_drop")
for (i in seq_along(cols))
    colData(sce)[[drop_cols[i]]] <- isOutlier(sce[[cols[i]]], 
        nmads = 3, type = type[i], log = log[i], batch = sce$sample_id)

# Overlap of outlier cells from two metrics
sapply(drop_cols, function(i) 
    sapply(drop_cols, function(j)
        sum(sce[[i]] & sce[[j]])))

colData(sce)$discard <- rowSums(data.frame(colData(sce)[,drop_cols])) > 0
table(colData(sce)$discard)

## Plot the metrics and highlight the discarded cells
plotColData(sce, x = "sample_id", y = "sum", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "detected", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "subsets_Mt_percent", 
            colour_by = "discard")
```


## MA plot of the two populations

We think that the second cell population in sample 5 with the lower number of UMIs and detected genes consists of damaged cells caused by the dissociation. Sample 5 and 6 are technical replicates and we thus expect similar distributions of the QC metrics. We want to know if there are any genes enriched in the second cell population or if we can safely remove the cells?

To check if we would discarded an specific cell identity with out stringent filtering, we compare the gene expression in the two cell populations (excluding the outlier cells as defined above).
If the second cell population is enriched for a specific cell identity, the corresponding marker genes will have high log2FC in the MA plot. Mitochondrial genes are in blue.

```{r, results='asis'}
for (i in c("5NC96", "6NC96")) {
cat("### ", i, "\n")
## all retained cells
part <- sce[,colData(sce)$sample_id == i & colData(sce)$discard == FALSE]
## second population: cells with less than 7000 detected genes in sample 5 
colData(part)$second_pop <- (colData(part)$sum < 7000 | 
  colData(part)$detected < 3000) 
plotColData(part, x = "sample_id", y = "sum", colour_by = "second_pop") + 
  scale_y_log10()
plotColData(part, x = "sample_id", y = "detected", colour_by = "second_pop") + 
  scale_y_log10()

lost <- calculateAverage(counts(part)[,colData(part)$second_pop])
kept <- calculateAverage(counts(part)[,!colData(part)$second_pop])
logged <- cpm(cbind(lost, kept), log=TRUE, prior.count=2)
logFC <- logged[,1] - logged[,2]
abundance <- rowMeans(logged)

label <- rep("", length(logFC))
top <- order(logFC, decreasing = TRUE)[1:20]
label[top]  <- rowData(sce)$symbol[top]
df <- data.frame(abundance = abundance, logFC = logFC, label = label)
mito <- grep("MT-", rownames(sce), value = TRUE)
p <- ggplot(df, aes(x = abundance, y = logFC, label = label)) + 
  geom_point(size = 2.5) +
  geom_point(data = df[mito,], color = "dodgerblue", size = 2.5) + 
  xlab("Average count") + ylab("logFC (lost/kept)") + 
  theme_bw(base_size = 16) + 
  geom_text_repel()
print(p)
cat("\n\n")
}
```

We decided to additionally filter the cells in sample 5 and 6.

```{r}
## filter the cells with less than 5000 UMIs in sample 5 and 6
colData(sce) %>% data.frame() %>% 
  filter(sample_id %in% c("5NC96", "6NC96")) %>% 
  group_by(sample_id, discard) %>% 
  summarise(below_cutoff = sum(sum < 5000))
colData(sce)$manual_discard_sum <- colData(sce)$sum < 5000 & 
  colData(sce)$sample_id %in% c("5NC96", "6NC96")
## filter the cells with less than 3000 detected genes in sample 5 and 6
colData(sce) %>% data.frame() %>% 
  filter(sample_id %in% c("5NC96", "6NC96")) %>% 
  group_by(sample_id, discard) %>% 
  summarise(below_cutoff = sum(detected < 2500) )
colData(sce)$manual_discard_detected <- colData(sce)$detected < 2500 & 
  colData(sce)$sample_id %in% c("5NC96", "6NC96")
## highlight all manually discarded cells
colData(sce)$manual_discard <- colData(sce)$manual_discard_sum |
                                   colData(sce)$manual_discard_detected
plotColData(sce, x = "sample_id", y = "sum", colour_by = "manual_discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "detected", colour_by = "manual_discard") + 
  scale_y_log10()

## highlight all discarded cells
colData(sce)$discard <- colData(sce)$manual_discard |
                                   colData(sce)$discard
plotColData(sce, x = "sample_id", y = "detected", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "sum", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "subsets_Mt_percent", 
            colour_by = "discard")
```


Plot the library size against the number of detected genes before and after filtering.
```{r, fig.width=12, fig.height = 10}
cd <- colData(sce)
layout(matrix(1:12, nrow = 3, byrow = TRUE))
for (i in levels(sce$sample_id)) {
  tmp <- cd[cd$sample_id == i,]
  heatscatter(tmp$sum, tmp$detected, log = "xy", 
              main = paste0(i, "-unfiltered"), xlab = "total counts", 
              ylab = "detected genes")
  heatscatter(tmp$sum[!tmp$discard], tmp$detected[!tmp$discard], 
              log = "xy", main = paste0(i, "-filtered"), xlab = "total counts", 
              ylab = "detected genes")    
}
```


## MA plot of all discarded over retained cells

MA plot of all discarded cells (outlier cells or cells in second population in sample 5) over the retained cells. Mitochondrial genes in blue.

```{r}
lost <- calculateAverage(counts(sce)[,colData(sce)$discard])
kept <- calculateAverage(counts(sce)[,!colData(sce)$discard])

logged <- cpm(cbind(lost, kept), log=TRUE, prior.count=2)
logFC <- logged[,1] - logged[,2]
abundance <- rowMeans(logged)

## only label genes with high logFC
label <- rep("", length(logFC))
top <- order(logFC, decreasing = TRUE)[1:20]
label[top]  <- rowData(sce)$symbol[top]
df <- data.frame(abundance = abundance, logFC = logFC, label = label)
mito <- grep("MT-", rownames(sce), value = TRUE)
p <- ggplot(df, aes(x = abundance, y = logFC, label = label)) + 
  geom_point(size = 2.5) +
  geom_point(data = df[mito,], color = "dodgerblue", size = 2.5) + 
  xlab("Average count") + ylab("logFC (lost/kept)") + 
  theme_bw(base_size = 16) + 
  geom_text_repel()
print(p)
```

### MA plot per sample {.tabset}
```{r, results='asis'}
for (s in levels(colData(sce)$sample_id)) {
  cat("#### ", s, "\n")
  part <- sce[,colData(sce)$sample_id == s]
  lost <- calculateAverage(counts(part)[,colData(part)$discard])
  kept <- calculateAverage(counts(part)[,!colData(part)$discard])
  logged <- cpm(cbind(lost, kept), log=TRUE, prior.count=2)
  logFC <- logged[,1] - logged[,2]
  abundance <- rowMeans(logged)
  ## only label genes with high logFC
  label <- rep("", length(logFC))
  top <- order(logFC, decreasing = TRUE)[1:20]
  label[top]  <- rowData(part)$symbol[top]
  df <- data.frame(abundance = abundance, logFC = logFC, label = label)
  mito <- grep("MT-", rownames(part), value = TRUE)
  p <- ggplot(df, aes(x = abundance, y = logFC, label = label)) + 
    geom_point(size = 2.5) +
    geom_point(data = df[mito,], color = "dodgerblue", size = 2.5) + 
    xlab("Average count") + ylab("logFC (lost/kept)") + 
    theme_bw(base_size = 16) + 
    geom_text_repel()
  print(p) 
  cat("\n\n")
}
```

## Removal of outlier cells
We remove the outlier cells and filter the genes:

```{r}
## summary of the kept cells
nr <- table(cd$sample_id)
nr_fil <- table(cd$sample_id[!cd$discard])
print(rbind(
    unfiltered = nr, filtered = nr_fil, 
    "%" = round(nr_fil / nr * 100, digits = 0)))

## discard the outlier cells
dim(sce)
sce <- sce[,!cd$discard]
dim(sce)

## we filter genes and require > 1 count in at least 20 cells
sce <- sce[rowSums(counts(sce) > 1) >= 20, ]
dim(sce)
```


## Save data to RDS

```{r}
saveRDS(sce, file.path("output", "sce_03_filtering.rds"))
```

