---
title: "Clustering"
author: "Katharina Hembach"
date: "5/28/2020"
output: 
  html_document:
    toc: true,
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```

```{r}
# increase future's maximum allowed size of exported globals to 4GB 
# the default is 2GB
options(future.globals.maxSize = 4096 * 1024 ^ 2)
```


### Load packages
```{r, message = FALSE}
library(cowplot)
library(ggplot2)
library(Seurat)
library(SingleCellExperiment)
library(stringr)
```

## Load data

```{r}
sce <- readRDS(file.path("output", "sce_03_filtering.rds"))
```

## Normalization
```{r normalization, warning = FALSE}
# create SeuratObject
so <- CreateSeuratObject(
    counts = counts(sce),
    meta.data = data.frame(colData(sce)),
    project = "neural_cultures")

# split by sample
cells_by_sample <- split(colnames(sce), sce$sample_id)
so <- lapply(cells_by_sample, function(i) subset(so, cells = i))

## log normalize the data using a scaling factor of 10000
so <- lapply(so, NormalizeData, verbose = FALSE, scale.factor = 10000, 
             normalization.method = "LogNormalize")
```

```{r variable-features, warning = FALSE}
## Identify the top 2000 genes with high cell-to-cell variation
so <- lapply(so, FindVariableFeatures, nfeatures = 2000, 
    selection.method = "vst", verbose = FALSE)

## Plot variable features 
for (i in names(so)) {
  # Identify the 10 most highly variable genes
  top10 <- head(VariableFeatures(so[[i]]), 10)
  p <- VariableFeaturePlot(so[[i]])
  p <- LabelPoints(plot = p, points = top10, 
              labels = str_split(top10, "\\.", simplify = TRUE)[,2], 
              repel = TRUE)
  print(p)
}
```

```{r integration, warning = FALSE}
# find anchors & integrate
as <- FindIntegrationAnchors(so, verbose = FALSE)
so <- IntegrateData(anchorset = as, dims = seq_len(30), verbose = FALSE)

## We scale the data so that mean expression is 0 and variance is 1, across cells
## We also regress out the number of UMIs and the percentag of mitochondrial UMIs 
## per cell.
DefaultAssay(so) <- "integrated"
so <- ScaleData(so, verbose = FALSE, vars.to.regress = c("sum", 
                                                         "subsets_Mt_percent"))
```

## Dimension reduction 

We perform dimension reduction with t-SNE and UMAP based on PCA results.
```{r dimension-reduction, warning = FALSE}
so <- RunPCA(so, npcs = 30, verbose = FALSE)
so <- RunTSNE(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, do.fast = TRUE, verbose = FALSE)
so <- RunUMAP(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, verbose = FALSE)
```

### Plot PCA results

```{r, fig.width = 12, fig.height = 8}
# top genes that are associated with the first two PCs
VizDimLoadings(so, dims = 1:2, reduction = "pca")
```

```{r, fig.width = 10, fig.height = 8}
## PCA plot 
DimPlot(so, reduction = "pca", group.by = "sample_id")
```

```{r}
# elbow plot with the ranking of PCs based on the % of variance explained
ElbowPlot(so, ndims = 30)
```

```{r,  fig.width = 15, fig.height = 20}
## heatmaps of the top 20 PCs and the 500 most extreme cells for each compenent
DimHeatmap(so, dims = 1:20, cells = 500, balanced = TRUE, nfeatures = 20 )
```


## Clustering

We cluster the cells using the reduced PCA dimensions.

```{r clustering, warning = FALSE}
so <- FindNeighbors(so, reduction = "pca", dims = seq_len(20), verbose = FALSE)
for (res in c(0.1, 0.2, 0.4, 0.8, 1, 1.2, 2))
    so <- FindClusters(so, resolution = res, random.seed = 1, verbose = FALSE)
```


## Dimension reduction plots 

We plot the dimension reduction (DR) and color by sample, group and cluster ID
```{r dr-plots, fig.width = 10, fig.height = 14, warning = FALSE}
thm <- theme(aspect.ratio = 1, legend.position = "none")
ps <- lapply(c("sample_id", "group_id", "ident"), function(u) {
    p1 <- DimPlot(so, reduction = "tsne", group.by = u) + thm
    p2 <- DimPlot(so, reduction = "umap", group.by = u)
    lgd <- get_legend(p2)
    p2 <- p2 + thm
    list(p1, p2, lgd)
    plot_grid(p1, p2, lgd, nrow = 1,
        rel_widths = c(1, 1, 0.5))
})
plot_grid(plotlist = ps, ncol = 1)
```

## QC on DR plots {.tabset}

```{r DR-QC, results = "asis", fig.width = 12}
cs <- sample(colnames(so), 1e4) ## subsample cells
.plot_features <- function(so, dr, id) {
    FeaturePlot(so, cells = cs, features = id, reduction = dr, pt.size = 0.4, 
                cols = c("grey", "blue")) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
}
ids <- c("sum", "detected", "subsets_Mt_percent")
for (id in ids) {
    cat("## ", id, "\n")
    p1 <- .plot_features(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none") + ggtitle("tSNE")
    p2 <- .plot_features(so, "umap", id) + theme(legend.position = "none") + 
      ggtitle("UMAP")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```

## Save Seurat object to RDS

```{r}
saveRDS(so, file.path("output", "so_04_clustering.rds"))
```

