---
title: "Cell cycle scoring"
author: "Katharina Hembach"
date: "6/18/2020"
output: 
  html_document:
    toc: true,
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```


### Load packages

```{r load-libs, message = FALSE, warning = FALSE}
library(cowplot)
library(ggplot2)
library(RColorBrewer)
library(viridis)
library(scran)
library(Seurat)
library(SingleCellExperiment)
library(stringr)
library(RCurl)
library(BiocParallel)
library(dplyr)
```


# Load data & convert to SCE

```{r load-data}
so <- readRDS(file.path("output", "so_04_clustering.rds"))
sce <- as.SingleCellExperiment(so, assay = "RNA")
colData(sce) <- as.data.frame(colData(sce)) %>% 
    mutate_if(is.character, as.factor) %>% 
    DataFrame(row.names = colnames(sce))
so <- SetIdent(so, value = "integrated_snn_res.0.4")
so@meta.data$cluster_id <- Idents(so)
sce$cluster_id <- Idents(so)
```


# Cell cycle scoring with Seurat

We assign each cell a cell cycle scores and visualize them in the DR plots.
We use known G2/M and S phase markers that come with the `Seurat` package. The markers are anticorrelated and cells that to not express the markers should be in G1 phase.

We compute cell cycle phase:
```{r cell-cycle}
DefaultAssay(so) <- "RNA"
# A list of cell cycle markers, from Tirosh et al, 2015
cc_file <- getURL("https://raw.githubusercontent.com/hbc/tinyatlas/master/cell_cycle/Homo_sapiens.csv") 
cc_genes <- read.csv(text = cc_file)
# match the marker genes to the features
m <- match(cc_genes$geneID[cc_genes$phase == "S"], 
           str_split(rownames(GetAssayData(so)),
                     pattern = "\\.", simplify = TRUE)[,1])
s_genes <- rownames(GetAssayData(so))[m]
(s_genes <- s_genes[!is.na(s_genes)])
m <- match(cc_genes$geneID[cc_genes$phase == "G2/M"], 
           str_split(rownames(GetAssayData(so)), 
                     pattern = "\\.", simplify = TRUE)[,1])
g2m_genes <- rownames(GetAssayData(so))[m]
(g2m_genes <- g2m_genes[!is.na(g2m_genes)])
so <- CellCycleScoring(so, s.features = s_genes, g2m.features = g2m_genes,
                       set.ident = TRUE)
DefaultAssay(so) <- "integrated"
```

# Cell cycle assignment using `cyclone`

```{r}
## read pretrained set of human cell cycle markers
human_pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", 
                                   package="scran"))
# Using Ensembl IDs to match up with the annotation in 'mm.pairs'.
assignments <- cyclone(sce, human_pairs, 
                       gene.names = str_split(rownames(sce), pattern = "\\.", 
                                              simplify = TRUE)[,1],
                       BPPARAM = MulticoreParam(workers = 20),
                       verbose = TRUE)
table(assignments$phases, colData(sce)$cluster_id)
## Add cell cycle phases to Seurat object
so$cyclone_phase <- assignments$phases
```

# Colored DR {.tabset}

```{r dim-red, results = "asis", fig.width = 12}
cs <- sample(colnames(so), 5e3)
.plot_dr <- function(so, dr, id)
    DimPlot(so, cells = cs, group.by = id, reduction = dr, pt.size = 0.4) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
ids <- c("cluster_id", "group_id", "sample_id", "Phase", "cyclone_phase")
for (id in ids) {
    cat("## ", id, "\n")
    p1 <- .plot_dr(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none")
    p2 <- .plot_dr(so, "umap", id) + theme(legend.position = "none")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```

## Save Seurat object to RDS

```{r}
saveRDS(so, file.path("output", "so_04_1_cell_cycle.rds"))
```
