---
title: "Cluster annotation"
author: "Katharina Hembach"
date: "5/29/2020"
output: 
  html_document:
    toc: true,
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```


### Load packages

```{r load-libs, message = FALSE, warning = FALSE}
library(ComplexHeatmap)
library(cowplot)
library(ggplot2)
library(dplyr)
library(muscat)
library(purrr)
library(RColorBrewer)
library(viridis)
library(scran)
library(Seurat)
library(SingleCellExperiment)
library(stringr)
```

# Load data & convert to SCE

```{r load-data}
so <- readRDS(file.path("output", "so_04_clustering.rds"))
sce <- as.SingleCellExperiment(so, assay = "RNA")
colData(sce) <- as.data.frame(colData(sce)) %>% 
    mutate_if(is.character, as.factor) %>% 
    DataFrame(row.names = colnames(sce))
```

# Nb. of clusters by resolution

```{r}
cluster_cols <- grep("res.[0-9]", colnames(colData(sce)), value = TRUE)
sapply(colData(sce)[cluster_cols], nlevels)
```

# Cluster-sample counts

```{r}
# set cluster IDs to resolution 0.4 clustering
so <- SetIdent(so, value = "integrated_snn_res.0.4")
so@meta.data$cluster_id <- Idents(so)
sce$cluster_id <- Idents(so)
(n_cells <- table(sce$cluster_id, sce$sample_id))
```

# Relative cluster-abundances

```{r fig.width = 6, fig.height = 5}
fqs <- prop.table(n_cells, margin = 2)
mat <- as.matrix(unclass(fqs))
Heatmap(mat,
    col = rev(brewer.pal(11, "RdGy")[-6]),
    name = "Frequency",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster_id",
    column_title = "sample_id",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(round(mat[j, i] * 100, 2), x = x, y = y, 
            gp = gpar(col = "white", fontsize = 8)))
```


# Cell cycle

We assign each cell a cell cycle scores and visualize them in the DR plots. 
We use known G2/M and S phase markers that come with the `Seurat` package. The markers are anticorrelated and cells that to not express the markers are supposed to be in G1 phase.
```{r}
# A list of cell cycle markers, from Tirosh et al, 2015, is loaded with Seurat.  We can
# segregate this list into markers of G2/M phase and markers of S phase
(s_genes <- cc.genes$s.genes)
(g2m_genes <- cc.genes$g2m.genes)
# match the marker genes to the features
m <- match(s_genes, str_split(rownames(GetAssayData(so, assay = "RNA")), 
                              pattern = "\\.", simplify = TRUE)[,2])
s_genes <- rownames(GetAssayData(so, assay = "RNA"))[m]
s_genes <- s_genes[!is.na(s_genes)] 
m <- match(g2m_genes, str_split(rownames(GetAssayData(so, assay = "RNA")), 
                              pattern = "\\.", simplify = TRUE)[,2])
g2m_genes <- rownames(GetAssayData(so, assay = "RNA"))[m]
g2m_genes <- g2m_genes[!is.na(g2m_genes)] 

# change the default assay to the raw counts
DefaultAssay(so) <- "RNA"
so <- CellCycleScoring(so, s.features = s_genes, g2m.features = g2m_genes, 
                       set.ident = TRUE)
# change back to integrated
DefaultAssay(so) <- "integrated"
```


# DR colored by cluster ID {.tabset}

```{r dim-red, results = "asis", fig.width = 12}
cs <- sample(colnames(so), 5e3)
.plot_dr <- function(so, dr, id)
    DimPlot(so, cells = cs, group.by = id, reduction = dr, pt.size = 0.4) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
ids <- c("cluster_id", "group_id", "sample_id", "Phase")
for (id in ids) {
    cat("## ", id, "\n")
    p1 <- .plot_dr(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none")
    p2 <- .plot_dr(so, "umap", id) + theme(legend.position = "none")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```

# Find markers using `scran`

We identify candidate marker genes for each cluster that enable a separation of that group from all other groups.

```{r scran-findMarkers}
scran_markers <- findMarkers(sce, 
    groups = sce$cluster_id, block = sce$sample_id, 
    direction = "up", lfc = 2, full.stats = TRUE)
```

## Heatmap of mean marker-exprs. by cluster
We aggregate the cells to pseudobulks and plot the average expression of the condidate marker genes in each of the clusters. 
```{r scran-markers-heatmap, fig.width = 12, fig.height = 6}
gs <- lapply(scran_markers, function(u) rownames(u)[u$Top == 1])
## candidate cluster markers
lapply(gs, function(x) str_split(x, pattern = "\\.", simplify = TRUE)[,2])
sub <- sce[unique(unlist(gs)), ]
pbs <- aggregateData(sub, assay = "logcounts", by = "cluster_id", fun = "mean")
mat <- t(muscat:::.scale(assay(pbs)))
## remove the Ensembl ID from the gene names
colnames(mat) <- str_split(colnames(mat), pattern = "\\.", simplify = TRUE)[,2] 
Heatmap(mat,
    name = "scaled avg.\nexpression",
    col = viridis(10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster_id",
    rect_gp = gpar(col = "white"))
```



