---
title: "Cell hashing test quality control"
author: "Katharina Hembach"
date: "5/14/2021"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")),
                      message = FALSE)
```

### Load packages
```{r, message = FALSE}
library(DropletUtils)
library(BiocParallel)
library(ggplot2)
library(scater)
library(readxl)
library(Seurat)
library(scales)
library(viridis)
library(dplyr)
```

# Importing CellRanger output and metadata
```{r import}
fs <- file.path("data", "cell_hashing_test", 
                "CellRangerCount_57443_2021-05-12--11-37-28", "HashTag_test", 
                "filtered_feature_bc_matrix.h5")
names(fs) <- "cell_hashing_test"
sce_raw <- read10xCounts(samples = fs)

# rename colnames and dimnames
rowData(sce_raw)$Type <- NULL
names(rowData(sce_raw)) <- c("ensembl_id", "symbol")
names(colData(sce_raw)) <- c("sample_id", "barcode")
sce_raw$sample_id <- factor(sce_raw$sample_id)

# load metadata
meta <- read_excel(file.path("data", "cell_hashing_test", "SampleName_feature_ref_MHP.xlsm"))
m <- match(meta$name, rowData(sce_raw)$symbol)

## separate gene counts from HTO counts
rowData(sce_raw) %>% tail
sce <- sce_raw[-m,]
dimnames(sce) <- list(with(rowData(sce), paste(ensembl_id, symbol, sep = ".")),
                      with(colData(sce), paste(barcode, sample_id, sep = ".")))
```


# Quality control
We compute cell-level QC.
```{r QC, message = FALSE}
# remove empty rows
sce <- sce[rowSums(counts(sce) > 0) > 0, ]
dim(sce)

(mito <- grep("MT-", rownames(sce), value = TRUE))
sce <- addPerCellQC(sce, subsets = list(Mt = mito))
# we compute the fraction of mitochondrial genes and the logit of it 
sce$subsets_Mt_fraction <- (sce$subsets_Mt_percent + 0.001) /100
sce$subsets_Mt_fraction_logit <- qlogis(sce$subsets_Mt_fraction + 0.001)
# library size
summary(sce$sum)
# number of detected genes per cell
summary(sce$detected)
# percentage of counts that come from mitochondrial genes:
summary(sce$subsets_Mt_percent)
```

## Diagnostic plots

The number of counts per cell:
```{r}
plotColData(sce, x = "sample_id", y = "sum") + scale_y_log10()
```

The number of genes:
```{r}
plotColData(sce, x = "sample_id", y = "detected") + scale_y_log10() 
```

The percentage of mitochondrial genes:
```{r}
plotColData(sce, x = "sample_id", y = "subsets_Mt_percent")
```

We plot the total number of counts against the number of detected genes and color by the fraction of mitochondrial genes:
```{r}
cd <- data.frame(colData(sce))
ggplot(cd, aes(x = sum, y = detected, color = subsets_Mt_fraction)) +
  geom_point(alpha = 0.7) + 
  geom_density_2d(color = "grey", bins = 6) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~sample_id) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("sum of counts") + 
  ylab("number of detected genes") + 
  labs(color = "mitochondrial fraction") +
  scale_color_viridis(trans = "logit", breaks = c(0.01, 0.1, 0.25, 0.5, 0.75))
```

We plot the total number of counts against the mitochondrial content. Well-behaved cells should have many expressed genes and a low fraction of mitochondrial genes. High mitochondrial content indicates empty or damaged cells. 
```{r}
ggplot(cd, aes(x = sum, y = subsets_Mt_fraction)) +
  geom_point(color = "darkgrey", alpha = 0.3) + 
  geom_density_2d(color = "lightblue") +
  scale_x_log10() +
  scale_y_continuous(trans = 'logit', 
                     breaks = c(0.01, 0.05, 0.1, 0.2, 0.5, 0.75)) +
  facet_wrap(~sample_id) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("sum of counts") + 
  ylab("logit(mitochondrial fraction)")
```


# Create Seurat object and split gene and HTO counts
```{r create-so}
## convert DDelayedMatrix to dgCMatrix for import into Seurat object
counts <- as(counts(sce, withDimnames = FALSE), "dgCMatrix")
colnames(counts) <- colnames(counts(sce))
rownames(counts) <- rownames(counts(sce))
                             
so <- CreateSeuratObject(
    counts = counts,
    meta.data = data.frame(colData(sce)),
    project = "cell_hashing_test")
## add HTO data as independent assay
hto_counts <- as(counts(sce_raw, withDimnames = FALSE)[m,with(colData(sce), 
                                    paste(barcode, sample_id, sep = ".")) %in% 
                                colnames(sce)], "dgCMatrix")
colnames(hto_counts) <- colnames(sce)
rownames(hto_counts) <- rownames(sce_raw)[m]
so[["HTO"]] <- CreateAssayObject(counts = hto_counts)
```

# Data normalization
```{r normalize-data}
DefaultAssay(so) <- "RNA"
# Normalize RNA data with log normalization
so <- NormalizeData(so)
# Find and scale variable features
so <- FindVariableFeatures(so, selection.method = "mean.var.plot")
so <- ScaleData(so, features = VariableFeatures(so))

# Normalize HTO data, here we use centered log-ratio (CLR) transformation
so <- NormalizeData(so, assay = "HTO", normalization.method = "CLR")
```

# Demultiplex cells based on HTO enrichment

```{r demultiplex-hto, message = TRUE}
so <- HTODemux(so, assay = "HTO", positive.quantile = 0.99)
```

## Visualize results

```{r demultiplexing-viz}
# Global classification results
table(so$HTO_classification.global)

# Group cells based on the max HTO signal
Idents(so) <- "HTO_maxID"
# Group cells based on the max HTO signal
RidgePlot(so, assay = "HTO", features = rownames(so[["HTO"]]), ncol = 3)

# Visualize pairs of HTO signals to check mutual exclusivity in singlets
DefaultAssay(object = so) <- "HTO"
FeatureScatter(so, feature1 = "B0253", feature2 = "B0254")
FeatureScatter(so, feature1 = "B0257", feature2 = "B0254")
FeatureScatter(so, feature1 = "B0253", feature2 = "B0257")

## compare number of UMIs for singlet's, doublets and negative cells
Idents(so) <- "HTO_classification.global"
VlnPlot(so, features = "nCount_HTO", pt.size = 0.1, log = TRUE)
VlnPlot(so, features = "nCount_RNA", pt.size = 0.1, log = TRUE)

## tSNE for HTOs
# First, we will remove negative cells from the object
subs <- subset(so, idents = "Negative", invert = TRUE)
subs$HTO_classification %>% table
# Calculate a tSNE embedding of the HTO data
DefaultAssay(subs) <- "HTO"
subs <- ScaleData(subs, features = rownames(subs), 
    verbose = FALSE)
subs <- RunPCA(subs, features = rownames(subs), approx = FALSE, npcs = 3)
subs <- RunTSNE(subs, dims = 1:3, perplexity = 100)
Idents(subs) <- "HTO_classification.global"
DimPlot(subs)
Idents(subs) <- 'HTO_classification'
DimPlot(subs)

# HTO heatmap
HTOHeatmap(so, assay = "HTO", ncells = 5000)
```


## Cluster based on gene counts and visualize cells
```{r gene-count-clustering}
DefaultAssay(so) <- "RNA"
# Extract the singlets
singlets <- subset(so, idents = "Singlet")
singlets$HTO_classification %>% table
# Select the top 1000 most variable features
singlets <- FindVariableFeatures(singlets, selection.method = "mean.var.plot")
# Scaling RNA data, we only scale the variable features here for efficiency
singlets <- ScaleData(singlets, features = VariableFeatures(singlets))
# Run PCA
singlets <- RunPCA(singlets, features = VariableFeatures(singlets))

# We select the top 10 PCs for clustering and tSNE based on PCElbowPlot
singlets <- FindNeighbors(singlets, reduction = "pca", dims = 1:10)
singlets <- FindClusters(singlets, resolution = 0.6, verbose = FALSE)
singlets <- RunTSNE(singlets, reduction = "pca", dims = 1:10)
singlets <- RunUMAP(singlets, reduction = "pca", dims = 1:10)

# Projecting singlet identities on TSNE visualization
DimPlot(singlets, group.by = "HTO_classification", reduction = "tsne")
DimPlot(singlets, group.by = "HTO_classification", reduction = "umap")
```



## Identification of low quality cells

Based on the QC metrics, we identify cells with lower quality:

```{r outlier-identification}
cols <- c("sum", "detected", "subsets_Mt_percent")
log <- c(TRUE, TRUE, FALSE)
type <- c("lower", "lower", "higher")

drop_cols <- paste0(cols, "_drop")
for (i in seq_along(cols))
    colData(sce)[[drop_cols[i]]] <- isOutlier(sce[[cols[i]]], 
        nmads = 1, type = type[i], log = log[i], batch = sce$sample_id)

# Overlap of outlier cells from two metrics
sapply(drop_cols, function(i) 
    sapply(drop_cols, function(j)
        sum(sce[[i]] & sce[[j]])))

colData(sce)$discard <- rowSums(data.frame(colData(sce)[,drop_cols])) > 0
table(colData(sce)$discard)

## Plot the metrics and highlight the discarded cells
plotColData(sce, x = "sample_id", y = "sum", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "detected", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "subsets_Mt_percent", 
            colour_by = "discard")


## we manually filter filter the cells with less than 2000 UMIs
colData(sce)$manual_discard_sum <- colData(sce)$sum < 1000 
## filter the cells with less than 800 detected genes
colData(sce)$manual_discard_detected <- colData(sce)$detected < 800 


## highlight all manually discarded cells
colData(sce)$manual_discard <- colData(sce)$manual_discard_sum |
                                   colData(sce)$manual_discard_detected
plotColData(sce, x = "sample_id", y = "sum", colour_by = "manual_discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "detected", colour_by = "manual_discard") + 
  scale_y_log10()

## highlight all discarded cells
colData(sce)$discard <- colData(sce)$manual_discard |
                                   colData(sce)$discard
plotColData(sce, x = "sample_id", y = "detected", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "sum", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "subsets_Mt_percent", 
            colour_by = "discard")
table(colData(sce)$discard)
```

We discard the outlier cells
```{r discard-outliers}
dim(sce)
sce_filtered <- sce[,!sce$discard]
dim(sce_filtered)
```


# Create Seurat object and split gene and HTO counts
```{r create-so-filtered}
## convert DDelayedMatrix to dgCMatrix for import into Seurat object
counts <- as(counts(sce_filtered, withDimnames = FALSE), "dgCMatrix")
colnames(counts) <- colnames(counts(sce_filtered))
rownames(counts) <- rownames(counts(sce_filtered))
                             
so_filtered <- CreateSeuratObject(
    counts = counts,
    meta.data = data.frame(colData(sce_filtered)),
    project = "cell_hashing_test")
## add HTO data as independent assay
hto_counts <- as(counts(sce_raw, withDimnames = FALSE)[m,with(colData(sce), 
                                    paste(barcode, sample_id, sep = ".")) %in% 
                                colnames(sce_filtered)], "dgCMatrix")
colnames(hto_counts) <- colnames(sce_filtered)
rownames(hto_counts) <- rownames(sce_raw)[m]
so_filtered[["HTO"]] <- CreateAssayObject(counts = hto_counts)
```

# Data normalization
```{r normalize-data-filtered}
DefaultAssay(so_filtered) <- "RNA"
# Normalize RNA data with log normalization
so_filtered <- NormalizeData(so_filtered)
# Find and scale variable features
so_filtered <- FindVariableFeatures(so_filtered, selection.method = "mean.var.plot")
so_filtered <- ScaleData(so_filtered, features = VariableFeatures(so_filtered))

# Normalize HTO data, here we use centered log-ratio (CLR) transformation
so_filtered <- NormalizeData(so_filtered, assay = "HTO", normalization.method = "CLR")
```


# Demultiplex cells based on HTO enrichment

```{r demultiplex-hto-filtered, message = TRUE}
so_filtered <- HTODemux(so_filtered, assay = "HTO", positive.quantile = 0.99)
```

## Visualize results

```{r demultiplexing-viz-filtered}
# Global classification results
table(so_filtered$HTO_classification.global)

# Group cells based on the max HTO signal
Idents(so_filtered) <- "HTO_maxID"
# Group cells based on the max HTO signal
RidgePlot(so_filtered, assay = "HTO", features = rownames(so_filtered[["HTO"]]), ncol = 3)

# Visualize pairs of HTO signals to check mutual exclusivity in singlets
DefaultAssay(object = so_filtered) <- "HTO"
FeatureScatter(so_filtered, feature1 = "B0253", feature2 = "B0254")
FeatureScatter(so_filtered, feature1 = "B0257", feature2 = "B0254")
FeatureScatter(so_filtered, feature1 = "B0253", feature2 = "B0257")

## compare number of UMIs for singlet's, doublets and negative cells
Idents(so_filtered) <- "HTO_classification.global"
VlnPlot(so_filtered, features = "nCount_HTO", pt.size = 0.1, log = TRUE)
VlnPlot(so_filtered, features = "nCount_RNA", pt.size = 0.1, log = TRUE)

## tSNE for HTOs
# First, we will remove negative cells from the object
subs <- subset(so_filtered, idents = "Negative", invert = TRUE)
subs$HTO_classification %>% table
# Calculate a tSNE embedding of the HTO data
DefaultAssay(subs) <- "HTO"
subs <- ScaleData(subs, features = rownames(subs), 
    verbose = FALSE)
subs <- RunPCA(subs, features = rownames(subs), approx = FALSE, npcs = 3)
subs <- RunTSNE(subs, dims = 1:3, perplexity = 100)
Idents(subs) <- "HTO_classification.global"
DimPlot(subs)
Idents(subs) <- 'HTO_classification'
DimPlot(subs)

# HTO heatmap
HTOHeatmap(so_filtered, assay = "HTO", ncells = 5000)
```


## Cluster based on gene counts and visualize cells
```{r gene-count-clustering-filtered}
DefaultAssay(so_filtered) <- "RNA"
# Extract the singlets
singlets_filtered <- subset(so_filtered, idents = "Singlet")
singlets_filtered$HTO_classification %>% table
# Select the top 1000 most variable features
singlets_filtered <- FindVariableFeatures(singlets_filtered, selection.method = "mean.var.plot")
# Scaling RNA data, we only scale the variable features here for efficiency
singlets_filtered <- ScaleData(singlets_filtered, features = VariableFeatures(singlets_filtered))
# Run PCA
singlets_filtered <- RunPCA(singlets_filtered, features = VariableFeatures(singlets_filtered))

# We select the top 10 PCs for clustering and tSNE based on PCElbowPlot
singlets_filtered <- FindNeighbors(singlets_filtered, reduction = "pca", dims = 1:10)
singlets_filtered <- FindClusters(singlets_filtered, resolution = 0.6, verbose = FALSE)
singlets_filtered <- RunTSNE(singlets_filtered, reduction = "pca", dims = 1:10)
singlets_filtered <- RunUMAP(singlets_filtered, reduction = "pca", dims = 1:10)

# Projecting singlet identities on TSNE visualization
DimPlot(singlets_filtered, group.by = "HTO_classification", reduction = "tsne")
DimPlot(singlets_filtered, group.by = "HTO_classification", reduction = "umap")
```


## Save data to RDS
```{r save-data}
saveRDS(sce, file.path("output", "CH-test-01-preprocessing.rds"))
saveRDS(singlets, file.path("output", "CH-test-01-preprocessing_singlets.rds"))
saveRDS(so, file.path("output", "CH-test-01-preprocessing_so.rds"))
saveRDS(singlets_filtered, file.path("output", "CH-test-01-preprocessing_singlets_filtered.rds"))
saveRDS(so_filtered, file.path("output", "CH-test-01-preprocessing_so_filtered.rds"))
```
