---
()title: "Cluster analysis of cell hashing test experiment"
author: "Katharina Hembach"
date: "5/19/2021"
output: 
  html_document:
    toc: true,
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```


### Load packages

```{r load-libs, message = FALSE, warning = FALSE}
library(ComplexHeatmap)
library(cowplot)
library(ggplot2)
library(dplyr)
library(muscat)
library(purrr)
library(RColorBrewer)
library(viridis)
library(scran)
library(Seurat)
library(SingleCellExperiment)
library(stringr)
```

# Load data & convert to SCE

```{r load-data}
so <- readRDS(file.path("output", "so_CH-test-02-transgene_expression.rds"))
sce <- as.SingleCellExperiment(so, assay = "RNA")
colData(sce) <- as.data.frame(colData(sce)) %>% 
    mutate_if(is.character, as.factor) %>% 
    DataFrame(row.names = colnames(sce))
```

# Number of clusters by resolution

```{r}
cluster_cols <- grep("res.[0-9]", colnames(colData(sce)), value = TRUE)
sapply(colData(sce)[cluster_cols], nlevels)
```

# Cluster-sample counts

```{r}
# set cluster IDs to resolution 0.4 clustering
so <- SetIdent(so, value = "RNA_snn_res.0.8")
so@meta.data$cluster_id <- Idents(so)
sce$cluster_id <- Idents(so)
(n_cells <- table(sce$cluster_id, sce$sample_id))
```

# Relative cluster-abundances

```{r fig.width = 6, fig.height = 5}
fqs <- prop.table(n_cells, margin = 2)
mat <- round(as.matrix(unclass(fqs))*100, 2)
colfunc <- colorRampPalette(c("ghostwhite", "deepskyblue4"))
Heatmap(mat,
    # col = rev(brewer.pal(11, "RdGy")[-6]),
    col = colfunc(10),
    name = "Percentage\nof cells",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster ID",
    column_title = "sample ID",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(mat[j, i], x = x, y = y, 
            gp = gpar(col = "black", fontsize = 10)))
```


# DR colored by cluster ID {.tabset}

```{r dim-red, results = "asis", fig.width = 12}
.plot_dr <- function(so, dr, id)
    DimPlot(so, group.by = id, reduction = dr, pt.size = 0.4) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
ids <- c("cluster_id", "sample_id")
for (id in ids) {
    cat("## ", id, "\n")
    p1 <- .plot_dr(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none")
    p2 <- .plot_dr(so, "umap", id) + theme(legend.position = "none")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```

# Find markers using `scran`

## Markers 
We identify candidate marker genes for each cluster that enable a separation of that group from any other group. The null hypothesis is that the log FC between a cluster and the compared cluster is 2.

```{r scran-findMarkers}
scran_markers <- findMarkers(sce, 
    groups = sce$cluster_id, block = sce$sample_id, 
    direction = "up", lfc = 2, full.stats = TRUE, log.p = FALSE)
```

We write tables with the top marker genes  per cluster.

```{r write-scran-markers}
# gs2 <- lapply(scran_markers, function(u) u[u$Top %in% 1:2,])
# for (i in seq_along(gs2)) {
#   write.table(x = gs2[[i]] %>% as.data.frame %>% 
#                 dplyr::mutate(gene = rownames(gs2[[i]])) %>%
#                 dplyr::relocate(gene), 
#               file =  file.path("output", 
#                                 paste0("TDP-06-no_integration_cluster", i-1, "_marker_genes.txt")),
#               sep = "\t", quote = FALSE, row.names = FALSE)
# }
```


## Heatmap of mean marker-exprs. by cluster
We aggregate the cells to pseudobulks and plot the average expression of the condidate marker genes in each of the clusters. 
```{r scran-markers-heatmap, fig.width = 12, fig.height = 6}
## including marker genes of rank 1 and 2
gs <- lapply(scran_markers, function(u) rownames(u)[u$Top %in% 1:2])
## candidate cluster markers
lapply(gs, function(x) {
  y <- str_split(x, pattern = "\\.", simplify = TRUE)[,2]
  y[which(y == "")] <- x[which(y == "")]
  y
  })

sub <- sce[unique(unlist(gs)), ]
pbs <- aggregateData(sub, assay = "logcounts", by = "cluster_id", fun = "mean")
mat <- t(muscat:::.scale(assay(pbs)))
## remove the Ensembl ID from the gene names
cnames <- colnames(mat)
colnames(mat) <- str_split(cnames, pattern = "\\.", simplify = TRUE)[,2] 
colnames(mat)[which(colnames(mat) == "")] <- cnames[which(colnames(mat) == "")]

Heatmap(mat,
    name = "scaled avg.\nexpression",
    col = viridis(10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster_id",
    rect_gp = gpar(col = "white"))
```


# Known marker genes

Apart from the usual marker genes, we also want to analyse the expression of Casein Kinase 1 Epsilon (CSNK1E).

```{r}
## source file with list of known marker genes
source(file.path("data", "known_cell_type_markers.R"))
fs[["kinase"]] <- "CSNK1E"

fs <- lapply(fs, sapply, function(g)
    grep(pattern = paste0("\\.", g, "$"), rownames(sce), value = TRUE)
  )

fs <- lapply(fs, function(x) unlist(x[lengths(x) !=0]) )

gs <- gsub(".*\\.", "", unlist(fs))
ns <- vapply(fs, length, numeric(1))
ks <- rep.int(names(fs), ns)
labs <- lapply(fs, function(x) gsub(".*\\.", "",x))
```

## Heatmap of mean marker-exprs. by cluster

```{r known-markers-heatmap, fig.width = 10, fig.height = 18}
# split cells by cluster
cs_by_k <- split(colnames(sce), sce$cluster_id)
# compute cluster-marker means
ms_by_cluster <- lapply(fs, function(gs) vapply(cs_by_k, function(i)
        Matrix::rowMeans(logcounts(sce)[gs, i, drop = FALSE]), 
        numeric(length(gs))))
# prep. for plotting & scale b/w 0 and 1
mat <- do.call("rbind", ms_by_cluster)
mat <- muscat:::.scale(mat)
rownames(mat) <- gs
cols <- muscat:::.cluster_colors[seq_along(fs)]
cols <- setNames(cols, names(fs))
row_anno <- rowAnnotation(
    df = data.frame(label = factor(ks, levels = names(fs))),
    col = list(label = cols), gp = gpar(col = "white"))
# percentage of cells from each of the samples per cluster
sample_props <- prop.table(n_cells, margin = 1)
col_mat <- as.matrix(unclass(sample_props))
sample_cols <- c("#882255", "#11588A",  "#117733")
sample_cols <- setNames(sample_cols, colnames(col_mat))
col_anno <- HeatmapAnnotation(
    perc_sample = anno_barplot(col_mat, gp = gpar(fill = sample_cols), 
                               height = unit(2, "cm"),
                               border = FALSE),
    annotation_label = "fraction of sample\nin cluster",
    gap = unit(10, "points"))
col_lgd <- Legend(labels = names(sample_cols),
       title = "sample",
       legend_gp = gpar(fill = sample_cols))

hm <- Heatmap(mat,
    name = "scaled avg.\nexpression",
    col = viridis(10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    column_title = "cluster_id",
    column_title_side = "bottom",
    column_names_side = "bottom",
    column_names_rot = 0, 
    column_names_centered = TRUE,
    rect_gp = gpar(col = "white"),
    left_annotation = row_anno,
    top_annotation = col_anno)
draw(hm, annotation_legend_list = list(col_lgd))
```


# DR colored by marker expression {.tabset}

```{r dim-red-expr, results = "asis", fig.width = 12, fig.height = 12}
# UMAPs colored by marker-expression
for (m in seq_along(fs)) {
    cat("## ", names(fs)[m], "\n")
    ps <- lapply(seq_along(fs[[m]]), function(i) {
        if (!fs[[m]][i] %in% rownames(so)) return(NULL)
        FeaturePlot(so, features = fs[[m]][i], reduction = "umap", 
                    pt.size = 0.4) +
            theme(aspect.ratio = 1, legend.position = "none") +
            ggtitle(labs[[m]][i]) + theme_void() + theme(aspect.ratio = 1)
    })
    # arrange plots in grid
    ps <- ps[!vapply(ps, is.null, logical(1))]
    p <- plot_grid(plotlist = ps, ncol = 4, label_size = 10)
    print(p)
    cat("\n\n")
}

## plot the expression of the endogenous TDP-43 and TDP-HA
transg <- c("ENSG00000120948.TARDBP", "ENSG00000120948.TARDBP-alevin", "TDP43-HA", 
         "GA50-EGFP", "HA-GA50")
names(transg) <- c("TARDBP", "TARDBP-alevin", "TDP-HA", "GA50-EGFP", "HA-GA50")
cat("## transgenes\n")
ps <- lapply(seq_along(transg), function(i) {
    if (!transg[i] %in% rownames(so)) return(NULL)
    FeaturePlot(so, features = transg[i], reduction = "umap", pt.size = 0.4) +
        theme(aspect.ratio = 1, legend.position = "none") +
        ggtitle(names(transg)[i]) + theme_void() + theme(aspect.ratio = 1)
})
# arrange plots in grid
ps <- ps[!vapply(ps, is.null, logical(1))]
p <- plot_grid(plotlist = ps, ncol = 4, label_size = 10)
print(p)
cat("\n\n")
```


# Reactive astrocyte markers

```{r read-astrocyte-markers}
## source file with list of known marker genes
source(file.path("data", "reactive_astrocyte_markers.R"))
fs <- lapply(fs, sapply, function(g)
    grep(pattern = paste0("\\.", g, "$"), rownames(sce), value = TRUE)
  )

fs <- lapply(fs, function(x) unlist(x[lengths(x) !=0]) )

gs <- gsub(".*\\.", "", unlist(fs))
ns <- vapply(fs, length, numeric(1))
ks <- rep.int(names(fs), ns)
labs <- lapply(fs, function(x) gsub(".*\\.", "",x))
```


## Heatmap of reactive astrocyte markers

```{r astrocyte-markers-heatmap, fig.width = 9, fig.height = 15}
# compute cluster-marker means
ms_by_cluster <- lapply(fs, function(gs) vapply(cs_by_k, function(i)
        Matrix::rowMeans(logcounts(sce)[gs, i, drop = FALSE]), 
        numeric(length(gs))))
# prep. for plotting & scale b/w 0 and 1
mat <- do.call("rbind", ms_by_cluster)
mat <- muscat:::.scale(mat)
rownames(mat) <- gs
cols <- muscat:::.cluster_colors[seq_along(fs)]
cols <- setNames(cols, names(fs))
row_anno <- rowAnnotation(
    df = data.frame(label = factor(ks, levels = names(fs))),
    col = list(label = cols), gp = gpar(col = "white"))
# percentage of cells from each of the samples per cluster
sample_props <- prop.table(n_cells, margin = 1)
col_mat <- as.matrix(unclass(sample_props))
sample_cols <- c("#882255", "#11588A",  "#117733")
sample_cols <- setNames(sample_cols, colnames(col_mat))
col_anno <- HeatmapAnnotation(
    perc_sample = anno_barplot(col_mat, gp = gpar(fill = sample_cols), 
                               height = unit(2, "cm"),
                               border = FALSE),
    annotation_label = "fraction of sample\nin cluster",
    gap = unit(10, "points"))
col_lgd <- Legend(labels = names(sample_cols),
       title = "sample",
       legend_gp = gpar(fill = sample_cols))

hm <- Heatmap(mat,
    name = "scaled avg.\nexpression",
    col = viridis(10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    column_title = "cluster_id",
    column_title_side = "bottom",
    column_names_side = "bottom",
    column_names_rot = 0, 
    column_names_centered = TRUE,
    rect_gp = gpar(col = "white"),
    left_annotation = row_anno,
    top_annotation = col_anno)
draw(hm, annotation_legend_list = list(col_lgd))
```


# Marker genes for virus cell tropism

```{r read-virus-markers}
## source file with list of known marker genes
source(file.path("data", "virus_cell_tropism_markers.R"))
fs <- lapply(fs, sapply, function(g)
    grep(pattern = paste0("\\.", g, "$"), rownames(sce), value = TRUE)
  )

fs <- lapply(fs, function(x) unlist(x[lengths(x) !=0]) )

gs <- gsub(".*\\.", "", unlist(fs))
ns <- vapply(fs, length, numeric(1))
ks <- rep.int(names(fs), ns)
labs <- lapply(fs, function(x) gsub(".*\\.", "",x))
```

## Heatmap of virus cell tropism markers

```{r virus-markers-heatmap, fig.width = 9, fig.height = 10}
# compute cluster-marker means
ms_by_cluster <- lapply(fs, function(gs) vapply(cs_by_k, function(i)
        Matrix::rowMeans(logcounts(sce)[gs, i, drop = FALSE]), 
        numeric(length(gs))))
# prep. for plotting & scale b/w 0 and 1
mat <- do.call("rbind", ms_by_cluster)
mat <- muscat:::.scale(mat)
rownames(mat) <- gs
cols <- muscat:::.cluster_colors[seq_along(fs)]
cols <- setNames(cols, names(fs))
row_anno <- rowAnnotation(
    df = data.frame(label = factor(ks, levels = names(fs))),
    col = list(label = cols), gp = gpar(col = "white"))
# percentage of cells from each of the samples per cluster
sample_props <- prop.table(n_cells, margin = 1)
col_mat <- as.matrix(unclass(sample_props))
sample_cols <- c("#882255", "#11588A",  "#117733")
sample_cols <- setNames(sample_cols, colnames(col_mat))
col_anno <- HeatmapAnnotation(
    perc_sample = anno_barplot(col_mat, gp = gpar(fill = sample_cols), 
                               height = unit(2, "cm"),
                               border = FALSE),
    annotation_label = "fraction of sample\nin cluster",
    gap = unit(10, "points"))
col_lgd <- Legend(labels = names(sample_cols),
       title = "sample",
       legend_gp = gpar(fill = sample_cols))

hm <- Heatmap(mat,
    name = "scaled avg.\nexpression",
    col = viridis(10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    column_title = "cluster_id",
    column_title_side = "bottom",
    column_names_side = "bottom",
    column_names_rot = 0, 
    column_names_centered = TRUE,
    rect_gp = gpar(col = "white"),
    left_annotation = row_anno,
    top_annotation = col_anno)
draw(hm, annotation_legend_list = list(col_lgd))
```

# Expression of cluster 12 markers genes (changed in TDP-HA expressing cells)

```{r cluster12-marker}
fs <- list(up = c("NPTX2", "FGF18", "TDP43-HA", "PCED1A", "MEF2A", "DYNC2H1", 
                   "APOE", "GADD45A", "BCAM", "DDIT3"),
           down = c("VGF", "SCG2", "GAP43", "C4orf48", "HINT1", "LY6H", 
                     "TUBA1A", "TMSB4X", "TUBB2B", "STMN2"))
fs <- lapply(fs, sapply, function(g)
    grep(pattern = paste0("\\.", g, "$"), rownames(sce), value = TRUE)
  )
fs[["up"]]["TDP43-HA"] <- "TDP43-HA"
fs <- lapply(fs, function(x) unlist(x[lengths(x) !=0]) )

gs <- gsub(".*\\.", "", unlist(fs))
ns <- vapply(fs, length, numeric(1))
ks <- rep.int(names(fs), ns)
labs <- lapply(fs, function(x) gsub(".*\\.", "",x))
```

## Heatmap of cluster 12 markers

```{r cluster12-markers-heatmap, fig.width = 9, fig.height = 10}
# compute cluster-marker means
ms_by_cluster <- lapply(fs, function(gs) vapply(cs_by_k, function(i)
        Matrix::rowMeans(logcounts(sce)[gs, i, drop = FALSE]), 
        numeric(length(gs))))
# prep. for plotting & scale b/w 0 and 1
mat <- do.call("rbind", ms_by_cluster)
mat <- muscat:::.scale(mat)
rownames(mat) <- gs
cols <- muscat:::.cluster_colors[seq_along(fs)]
cols <- setNames(cols, names(fs))
row_anno <- rowAnnotation(
    df = data.frame(label = factor(ks, levels = names(fs))),
    col = list(label = cols), gp = gpar(col = "white"))
# percentage of cells from each of the samples per cluster
sample_props <- prop.table(n_cells, margin = 1)
col_mat <- as.matrix(unclass(sample_props))
sample_cols <- c("#882255", "#11588A",  "#117733")
sample_cols <- setNames(sample_cols, colnames(col_mat))
col_anno <- HeatmapAnnotation(
    perc_sample = anno_barplot(col_mat, gp = gpar(fill = sample_cols), 
                               height = unit(2, "cm"),
                               border = FALSE),
    annotation_label = "fraction of sample\nin cluster",
    gap = unit(10, "points"))
col_lgd <- Legend(labels = names(sample_cols),
       title = "sample",
       legend_gp = gpar(fill = sample_cols))

hm <- Heatmap(mat,
    name = "scaled avg.\nexpression",
    col = viridis(10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    column_title = "cluster_id",
    column_title_side = "bottom",
    column_names_side = "bottom",
    column_names_rot = 0, 
    column_names_centered = TRUE,
    rect_gp = gpar(col = "white"),
    left_annotation = row_anno,
    top_annotation = col_anno)
draw(hm, annotation_legend_list = list(col_lgd))
```


## Save cluster markers to RDS

```{r save-rds}
saveRDS(scran_markers, file.path("output", "CH-test-03_scran_markers.rds"))
saveRDS(so, file.path("output", "CH-test-03-cluster-analysis_so.rds"))
```

