---
title: "Annotation of cell identity after Lam et al. NSC integration"
author: "Katharina Hembach"
date: "7/6/2020"
output: 
  html_document:
    toc: true,
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```

### Load packages

```{r load-libs, message = FALSE, warning = FALSE}
library(ComplexHeatmap)
library(cowplot)
library(ggplot2)
library(dplyr)
library(muscat)
library(purrr)
library(RColorBrewer)
library(viridis)
library(scran)
library(Seurat)
library(SingleCellExperiment)
library(stringr)
library(RCurl)
library(BiocParallel)
```

# Load data & convert to SCE

```{r load-data}
so <- readRDS(file.path("output", "Lam-01-clustering.rds"))
sce <- as.SingleCellExperiment(so, assay = "RNA")
colData(sce) <- as.data.frame(colData(sce)) %>% 
    mutate_if(is.character, as.factor) %>% 
    DataFrame(row.names = colnames(sce))
```

# Number of clusters by resolution

```{r}
cluster_cols <- grep("res.[0-9]", colnames(colData(sce)), value = TRUE)
sapply(colData(sce)[cluster_cols], nlevels)
```

# Cluster-sample counts

```{r}
# set cluster IDs to resolution 0.4 clustering
so <- SetIdent(so, value = "integrated_snn_res.0.4")
so@meta.data$cluster_id <- Idents(so)
sce$cluster_id <- Idents(so)
(n_cells <- table(sce$cluster_id, sce$sample_id))
```

# Relative cluster-abundances

```{r fig.width = 4, fig.height = 4}
fqs <- prop.table(n_cells, margin = 2)
mat <- as.matrix(unclass(fqs))
Heatmap(mat,
    col = rev(brewer.pal(11, "RdGy")[-6]),
    name = "Frequency",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster_id",
    column_title = "sample_id",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(round(mat[j, i] * 100, 2), x = x, y = y, 
            gp = gpar(col = "white", fontsize = 10)))
```


# Distribution of NES subtypes per cluster
In the paper, they identified clusters that were specific for different cell types. For our analysis, we merge identical cell subtypes from the different cell lines.
```{r fig.width = 5.5, fig.height = 5}
levels(sce$cell_subtype_nes) 
## merge identical cell subtypes
levels(sce$cell_subtype_nes)  <- 
  c("Glia_progenitor", "Neural_prog_Proliferating", "Neural_progenitor", 
    "Neural_progenitor", "Neural_progenitor", "Neuroblast", 
    "Radial_Glia_progenitor")
levels(sce$cell_subtype_nes) 
(n_types <- table(sce$cluster_id, sce$cell_subtype_nes))

fqs <- prop.table(n_types, margin = 2)
mat <- as.matrix(unclass(fqs))
Heatmap(mat,
    col = rev(brewer.pal(11, "RdGy")[-6]),
    name = "Frequency",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster_id",
    column_title = "sample_id",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(round(mat[j, i] * 100, 2), x = x, y = y, 
            gp = gpar(col = "white", fontsize = 10)))
```



# DR colored by cluster ID {.tabset}

```{r dim-red, results = "asis", fig.width = 12}
.plot_dr <- function(so, dr, id)
    DimPlot(so, group.by = id, reduction = dr, pt.size = 0.4) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
ids <- c("cluster_id", "group_id", "sample_id")
for (id in ids) {
    cat("## ", id, "\n")
    p1 <- .plot_dr(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none")
    p2 <- .plot_dr(so, "umap", id) + theme(legend.position = "none")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```


# Cluster markers from Lam et al. 

Similar to figure 2f in paper.

```{r}
## source file with list of known marker genes
source(file.path("data", "Lam_figure2_markers.R"))

fs <- lapply(fs, sapply, function(g)
    grep(pattern = paste0("\\.", g, "$"), rownames(sce), value = TRUE)
  )

fs <- lapply(fs, function(x) unlist(x[lengths(x) !=0]) )

gs <- gsub(".*\\.", "", unlist(fs))
ns <- vapply(fs, length, numeric(1))
ks <- rep.int(names(fs), ns)
labs <- lapply(fs, function(x) gsub(".*\\.", "",x))
```

## Heatmap of mean marker-exprs. by cluster

```{r known-markers-heatmap, fig.width = 10, fig.height = 18}
# split cells by cluster
cs_by_k <- split(colnames(sce), sce$cluster_id)
# compute cluster-marker means
ms_by_cluster <- lapply(fs, function(gs) vapply(cs_by_k, function(i)
        Matrix::rowMeans(logcounts(sce)[gs, i, drop = FALSE]), 
        numeric(length(gs))))
# prep. for plotting & scale b/w 0 and 1
mat <- do.call("rbind", ms_by_cluster)
mat <- muscat:::.scale(mat)
rownames(mat) <- gs
cols <- muscat:::.cluster_colors[seq_along(fs)]
cols <- setNames(cols, names(fs))
row_anno <- rowAnnotation(
    df = data.frame(label = factor(ks, levels = names(fs))),
    col = list(label = cols), gp = gpar(col = "white"))
# percentage of cells from each of the samples per cluster
sample_props <- prop.table(n_cells, margin = 1)
col_mat <- as.matrix(unclass(sample_props))
sample_cols <- c("#882255", "#CC6677", "#11588A")
sample_cols <- setNames(sample_cols, colnames(col_mat))
col_anno <- HeatmapAnnotation(
    perc_sample = anno_barplot(col_mat, gp = gpar(fill = sample_cols), 
                               height = unit(2, "cm"),
                               border = FALSE),
    annotation_label = "fraction of sample\nin cluster",
    gap = unit(10, "points"))
col_lgd <- Legend(labels = names(sample_cols),
       title = "sample",
       legend_gp = gpar(fill = sample_cols))

hm <- Heatmap(mat,
    name = "scaled avg.\nexpression",
    col = viridis(10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    column_title = "cluster_id",
    column_title_side = "bottom",
    column_names_side = "bottom",
    column_names_rot = 0, 
    column_names_centered = TRUE,
    rect_gp = gpar(col = "white"),
    left_annotation = row_anno,
    top_annotation = col_anno)
draw(hm, annotation_legend_list = list(col_lgd))
```


