---
title: "Annotation of cell identity after Lam et al. NSC integration"
author: "Katharina Hembach"
date: "7/6/2020"
output: 
  html_document:
    toc: true,
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```

### Load packages

```{r load-libs, message = FALSE, warning = FALSE}
library(ComplexHeatmap)
library(cowplot)
library(ggplot2)
library(dplyr)
library(muscat)
library(purrr)
library(RColorBrewer)
library(viridis)
library(scran)
library(Seurat)
library(SingleCellExperiment)
library(stringr)
library(RCurl)
library(BiocParallel)
library(CellMixS)
```

# Load data & convert to SCE

```{r load-data}
so <- readRDS(file.path("output", "Lam-01-clustering.rds"))
sce <- as.SingleCellExperiment(so, assay = "RNA")
colData(sce) <- as.data.frame(colData(sce)) %>% 
    mutate_if(is.character, as.factor) %>% 
    DataFrame(row.names = colnames(sce))
```

# Number of clusters by resolution

```{r}
cluster_cols <- grep("res.[0-9]", colnames(colData(sce)), value = TRUE)
sapply(colData(sce)[cluster_cols], nlevels)
```

# Cluster-sample counts

```{r}
# set cluster IDs to resolution 0.4 clustering
so <- SetIdent(so, value = "integrated_snn_res.0.4")
so@meta.data$cluster_id <- Idents(so)
sce$cluster_id <- Idents(so)
(n_cells <- table(sce$cluster_id, sce$sample_id))
```

# Relative cluster-abundances

```{r fig.width = 4, fig.height = 4}
fqs <- prop.table(n_cells, margin = 2)
mat <- round(as.matrix(unclass(fqs))*100, 2)
colfunc <- colorRampPalette(c("ghostwhite", "deepskyblue4"))
Heatmap(mat,
    col = colfunc(10),
    name = "Percentage\nof cells",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster ID",
    column_title = "sample ID",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(mat[j, i], x = x, y = y, 
            gp = gpar(col = "black", fontsize = 10)))
```

We split the cells from Lam et al. into the three different cell lines that they used in the paper.
```{r fig.width = 5, fig.height = 4}
ind <- which(sce$sample_id == "NES")
cell_label <- sce$sample_id
levels(cell_label) <- c(levels(cell_label), levels(sce$Cell_line))
cell_label[ind] <- sce$Cell_line[ind]
cell_label <- droplevels(cell_label)
levels(cell_label)[levels(cell_label)==".SAi2"] <- "SAi2"
so$cell_label <- cell_label

(n_cells_line <- table(sce$cluster_id, cell_label))
fqs <- prop.table(n_cells_line, margin = 2)
mat <- round(as.matrix(unclass(fqs))*100, 2)
Heatmap(mat,
    col = colfunc(10),
    name = "Percentage\nof cells",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster ID",
    column_title = "sample ID",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(mat[j, i], x = x, y = y, 
            gp = gpar(col = "black", fontsize = 10)))
```


# Distribution of NES subtypes per cluster
In the paper, they identified clusters that were specific for different cell types. For our analysis, we merge identical cell subtypes from the different cell lines.
```{r fig.width = 5.5, fig.height = 5}
levels(sce$cell_subtype_nes) 
## merge identical cell subtypes
levels(sce$cell_subtype_nes)  <- 
  c("Glia_progenitor", "Neural_prog_Proliferating", "Neural_progenitor", 
    "Neural_progenitor", "Neural_progenitor", "Neuroblast", 
    "Radial_Glia_progenitor")
levels(sce$cell_subtype_nes) 
(n_types <- table(sce$cluster_id, sce$cell_subtype_nes))

fqs <- prop.table(n_types, margin = 2)
mat <- round(as.matrix(unclass(fqs))*100, 2)
Heatmap(mat,
    col = colfunc(10),
    name = "Percentage\nof cells",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster ID",
    column_title = "sample ID",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(mat[j, i], x = x, y = y, 
            gp = gpar(col = "black", fontsize = 10)))
```



# DR colored by cluster ID {.tabset}

```{r dim-red, results = "asis", fig.width = 12}
.plot_dr <- function(so, dr, id)
    DimPlot(so, group.by = id, reduction = dr, pt.size = 0.4) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
ids <- c("cluster_id", "group_id", "sample_id", "cell_label")
for (id in ids) {
    cat("## ", id, "\n")
    p1 <- .plot_dr(so, "tsne", id)
    p2 <- .plot_dr(so, "umap", id)
    if(id == "group_id") {
      p1 <- p1 + scale_color_manual(values = c("springgreen3", "darkmagenta"))
      p2 <- p2 + scale_color_manual(values = c("springgreen3", "darkmagenta"))
    }
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none")
    p2 <- p2 + theme(legend.position = "none")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```


# Cluster markers from Lam et al. 

Similar to figure 2f in paper.

```{r}
## source file with list of known marker genes
source(file.path("data", "Lam_figure2_markers.R"))

fs <- lapply(fs, sapply, function(g)
    grep(pattern = paste0("\\.", g, "$"), rownames(sce), value = TRUE)
  )

fs <- lapply(fs, function(x) unlist(x[lengths(x) !=0]) )

gs <- gsub(".*\\.", "", unlist(fs))
ns <- vapply(fs, length, numeric(1))
ks <- rep.int(names(fs), ns)
labs <- lapply(fs, function(x) gsub(".*\\.", "",x))
```

## Heatmap of mean marker-exprs. by cluster

```{r known-markers-heatmap, fig.width = 10, fig.height = 18}
# split cells by cluster
cs_by_k <- split(colnames(sce), sce$cluster_id)
# compute cluster-marker means
ms_by_cluster <- lapply(fs, function(gs) vapply(cs_by_k, function(i)
        Matrix::rowMeans(logcounts(sce)[gs, i, drop = FALSE]), 
        numeric(length(gs))))
# prep. for plotting & scale b/w 0 and 1
mat <- do.call("rbind", ms_by_cluster)
mat <- muscat:::.scale(mat)
rownames(mat) <- gs
cols <- muscat:::.cluster_colors[seq_along(fs)]
cols <- setNames(cols, names(fs))
row_anno <- rowAnnotation(
    df = data.frame(label = factor(ks, levels = names(fs))),
    col = list(label = cols), gp = gpar(col = "white"))
# percentage of cells from each of the samples per cluster
sample_props <- prop.table(n_cells, margin = 1)
col_mat <- as.matrix(unclass(sample_props))
sample_cols <- c("#882255", "#CC6677", "#11588A")
sample_cols <- setNames(sample_cols, colnames(col_mat))
col_anno <- HeatmapAnnotation(
    perc_sample = anno_barplot(col_mat, gp = gpar(fill = sample_cols), 
                               height = unit(2, "cm"),
                               border = FALSE),
    annotation_label = "fraction of sample\nin cluster",
    gap = unit(10, "points"))
col_lgd <- Legend(labels = names(sample_cols),
       title = "sample",
       legend_gp = gpar(fill = sample_cols))

hm <- Heatmap(mat,
    name = "scaled avg.\nexpression",
    col = viridis(10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    column_title = "cluster_id",
    column_title_side = "bottom",
    column_names_side = "bottom",
    column_names_rot = 0, 
    column_names_centered = TRUE,
    rect_gp = gpar(col = "white"),
    left_annotation = row_anno,
    top_annotation = col_anno)
draw(hm, annotation_legend_list = list(col_lgd))
```


# Explore data integration with CellMixS

We use the [CellMixS](https://github.com/almutlue/CellMixS) Bioconductor R package to evaluate the data integration and potential batch effects. 
We test how well the two dataset are mixing or if there are batch effect with the Cellspecific Mixing Score (CMS), a test for batch effects within k-nearest neighbouring cells. 
A high cms score refers to good mixing, while a low score indicates batch-specific bias. The test considers differences in the number of cells from each batch.

```{r cellmixs-cms, fig.width = 12}
sce$group_id %>% table

## using PCA based on integrated and scaled data
## we set k high but below the size of the smallest group 
## because we want to evaluate global structures
## k_min is used to define the minimum size of the local neighbourhoods
sce <- cms(sce, k = 700, k_min = 200, group = "group_id", dim_red = "PCA",
           n_dim = 10, unbalanced = TRUE, 
           BPPARAM = MulticoreParam(workers = 15))
head(colData(sce)[,c("cms_smooth", "cms")])

## cms histogram
visHist(sce)
p1 <- visMetric(sce, metric_var = "cms_smooth", dim_red = "UMAP") + 
  theme_void() + theme(aspect.ratio = 1)
p2 <- visMetric(sce, metric_var = "cms", dim_red = "UMAP") + 
  theme_void() + theme(aspect.ratio = 1)
plot_grid(p1, p2)

## score distribution per cluster
p1 <- visCluster(sce, metric_var = "cms", cluster_var = "cluster_id") + 
  scale_fill_hue() +  
  scale_y_discrete(limits = rev(unique(sort(sce$cluster_id))))
p2 <- visCluster(sce, metric_var = "cms_smooth", cluster_var = "cluster_id") + 
  scale_fill_hue() + 
  scale_y_discrete(limits = rev(unique(sort(sce$cluster_id))))
plot_grid(p1, p2)
```


We also test how well the two datasets are integrated with the Local Density Differences (ldfDiff) metric.
In an optimal case relative densities (according to the same set of cells) should not change by integration and the ldfDiff score should be close to 0. In general the overall distribution of ldfDiff should be centered around 0 without long tails.
```{r cellmixs-integration-evaluation}
sce_int <- as.SingleCellExperiment(so, assay = "integrated")
colData(sce_int) <- as.data.frame(colData(sce_int)) %>% 
    mutate_if(is.character, as.factor) %>% 
    DataFrame(row.names = colnames(sce_int))

sce_pre_list <- list("P22" = sce[,sce$group_id == "P22"], 
                     "NES" = sce[,sce$group_id == "NES"])
## remove dimension reduction from integrated data
sce_pre_list <- lapply(sce_pre_list, function(x) {reducedDims(x) <- NULL; x})

sce_int <- ldfDiff(sce_pre_list, sce_combined = sce_int, group = "group_id",
               k = 7, dim_red = "PCA", dim_combined = "PCA", 
               assay_pre = "logcounts", assay_combined = "logcounts",
               n_dim = 3, res_name = "Seurat")

visIntegration(sce_int, metric = "diff_ldf", metric_name = "ldfDiff") 
```


