---
title: "Integrate Kanton et al. by groups"
author: "Katharina Hembach"
date: "8/26/2020"
ooutput: 
  html_document:
    toc: true,
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```

### Load packages
```{r, message = FALSE}
library(BiocParallel)
library(ggplot2)
library(dplyr)
library(cowplot)
library(ggplot2)
library(Seurat)
library(SingleCellExperiment)
library(future)
```


```{r future-setup}
# increase future's maximum allowed size of exported globals
# the default is 2GB
options(future.globals.maxSize = 11000 * 1024 ^ 2)
# change the current plan to access parallelization
plan("multiprocess", workers = 20)
```

## Load data & convert to SCE

```{r load-data}
so_nsc <- readRDS(file.path("output", "so_04_clustering.rds"))
DefaultAssay(so_nsc) <- "RNA"
```

```{r load-organoid-data}
so_org <- readRDS(file.path("output", "so_organoid-01-clustering.rds"))
DefaultAssay(so_org) <- "RNA"
```


## Integration

We integrate the two datasets using our three groups and the two cell lines from Kanton et al.

```{r integration-normalisation, warning = FALSE}
## make sure column names match between datasets:
## we will use sample_id, group_id, nUMI, fractionMt, nGene
so_nsc$nUMI <- so_nsc$sum
so_nsc$fractionMt <- so_nsc$subsets_Mt_fraction
so_nsc$nGene <- so_nsc$detected
so_nsc$sum <- NULL
so_nsc$subsets_Mt_fraction <- NULL
so_nsc$detected <- NULL
so_org$sample_id <- so_org$Line
so_org$group_id <- so_org$Stage
so_org$fractionMt <- so_org$PercentMito
so_org$Line <- NULL
so_org$Stage <- NULL
so_org$PercentMito <- NULL

so_nsc$integration_group <- so_nsc$group_id
so_org$integration_group <- so_org$sample_id

# split our cells by group
cells_by_sample <- split(colnames(so_nsc), so_nsc$integration_group)
so_nsc <- lapply(cells_by_sample, function(i) subset(so_nsc, cells = i))
# split organoid cells by cell Line
cells_by_sample <- split(colnames(so_org), so_org$sample_id)
so_org <- lapply(cells_by_sample, function(i) subset(so_org, cells = i))
## we combine the two lists
so <- c(so_nsc, so_org)

## Identify the top 2000 genes with high cell-to-cell variation
so <- lapply(so, FindVariableFeatures, nfeatures = 2000, 
    selection.method = "vst", verbose = FALSE)
```

```{r integration, warning = FALSE}
## find anchors & integrate
as <- FindIntegrationAnchors(so, verbose = FALSE)
so <- IntegrateData(anchorset = as, dims = seq_len(30), verbose = FALSE)
```

```{r integration-scale-data, warning = FALSE}
DefaultAssay(so) <- "integrated"
## We scale the data
so <- ScaleData(so, verbose = FALSE,
                vars.to.regress = c("nUMI", "fractionMt"))
```

```{r integration-DR, warning = FALSE}
so <- RunPCA(so, npcs = 30, verbose = FALSE)
so <- RunTSNE(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, do.fast = TRUE, verbose = FALSE)
so <- RunUMAP(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, verbose = FALSE)

## PCA plot 
DimPlot(so, reduction = "pca", group.by = "sample_id")

# elbow plot with the ranking of PCs based on the % of variance explained
ElbowPlot(so, ndims = 30)
```

```{r integration-clustering, warning = FALSE}
so <- FindNeighbors(so, reduction = "pca", dims = seq_len(20), verbose = FALSE)
so <- FindClusters(so, resolution = 0.4, random.seed = 1, verbose = FALSE)
```

```{r}
## we use factors for plotting
so$sample_id <- factor(so$sample_id, levels = c("1NSC", "2NSC", "3NC52", 
                                                "4NC52", "5NC96", "6NC96", 
                                                "H9", "409b2"))
## order levels according to experiment timeline (Fig. 1a)
so$group_id <- factor(so$group_id, levels = c("P22", "D52", "D96", "iPSCs", 
                                              "EB", "Neuroectoderm", 
                                              "Neuroepithelium", "Organoid-1M", 
                                              "Organoid-2M", "Organoid-4M"))
```


## Dimension reduction plots {.tabset} 

We plot the dimension reduction (DR) and color by the groups used for integration, sample/cell line, group/Stage, organoid cluster labels, cluster ID.

```{r integration-dim-red, results = "asis", fig.width = 12}
# set cluster IDs to resolution 0.4 clustering
so <- SetIdent(so, value = "integrated_snn_res.0.4")
so@meta.data$cluster_id <- Idents(so)

cs <- sample(colnames(so), 10e3)
.plot_dr <- function(so, dr, id)
    DimPlot(so, cells = cs, group.by = id, reduction = dr, pt.size = 0.4) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
ids <- c("integration_group", "sample_id", "group_id", "cl_FullLineage", "ident")
for (id in ids) {
    cat("### ", id, "\n")
    p1 <- .plot_dr(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none")
    p2 <- .plot_dr(so, "umap", id) + theme(legend.position = "none")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.5))
    print(p)
    cat("\n\n")
}
```


## QC on DR plots {.tabset}

```{r integration-DR-QC, results = "asis", fig.width = 12}
.plot_features <- function(so, dr, id) {
    FeaturePlot(so, cells = cs, features = id, reduction = dr, pt.size = 0.4, 
                cols = c("grey", "blue")) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
}
ids <- c("nUMI", "nGene", "fractionMt")
for (id in ids) {
    cat("### ", id, "\n")
    p1 <- .plot_features(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none") + ggtitle("tSNE")
    p2 <- .plot_features(so, "umap", id) + theme(legend.position = "none") + 
      ggtitle("UMAP")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```

## Save Seurat object to RDS

```{r save-so}
saveRDS(so, file.path("output", "so_04-group_integration.rds"))
```

