---
title: "Cluster analysis of group integrated cells"
author: "Katharina Hembach"
date: "8/26/2020"
output: 
  html_document:
    toc: true,
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```

### Load packages

```{r load-libs, message = FALSE, warning = FALSE}
library(ComplexHeatmap)
library(cowplot)
library(ggplot2)
library(dplyr)
library(muscat)
library(RColorBrewer)
library(Seurat)
library(SingleCellExperiment)
library(scran)
library(stringr)
library(viridis)
```

# Load data & convert to SCE

```{r load-data}
so <- readRDS(file.path("output", "so_04-group_integration.rds"))
sce <- as.SingleCellExperiment(so, assay = "RNA")
colData(sce) <- as.data.frame(colData(sce)) %>% 
    mutate_if(is.character, as.factor) %>% 
    DataFrame(row.names = colnames(sce))
```

# Cluster-sample counts

```{r}
# set cluster IDs to resolution 0.4 clustering
so <- SetIdent(so, value = "integrated_snn_res.0.4")
so@meta.data$cluster_id <- Idents(so)
sce$cluster_id <- Idents(so)
(n_cells <- table(sce$cluster_id, sce$sample_id))
```

# Relative cluster-abundances

```{r fig.width = 6, fig.height = 6}
fqs <- prop.table(n_cells, margin = 2)
mat <- as.matrix(unclass(fqs))
Heatmap(mat,
    col = rev(brewer.pal(11, "RdGy")[-6]),
    name = "Frequency",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster_id",
    column_title = "sample_id",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(round(mat[j, i] * 100, 2), x = x, y = y, 
            gp = gpar(col = "white", fontsize = 10)))
```

```{r fig.width = 9, fig.height = 4.5}
n_cells <- table(sce$sample_id, sce$cluster_id)
fqs <- prop.table(n_cells, margin = 2)
mat <- as.matrix(unclass(fqs))
Heatmap(mat,
    col = rev(brewer.pal(11, "RdGy")[-6]),
    name = "Frequency",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "sample_id",
    column_title = "cluster_id",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(round(mat[j, i] * 100, 2), x = x, y = y, 
            gp = gpar(col = "white", fontsize = 10)))

```

```{r fig.width = 6, fig.height = 6}
(n_cells_group <- table(sce$cluster_id, sce$group_id))
fqs <- prop.table(n_cells_group, margin = 2)
mat <- as.matrix(unclass(fqs))
Heatmap(mat,
    col = rev(brewer.pal(11, "RdGy")[-6]),
    name = "Frequency",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster_id",
    column_title = "group_id",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(round(mat[j, i] * 100, 2), x = x, y = y, 
            gp = gpar(col = "white", fontsize = 10)))
```

```{r fig.width = 10, fig.height = 8}
n_cells_lineage <- table(sce$cluster_id, sce$cl_FullLineage)
fqs <- prop.table(n_cells_lineage, margin = 2)
mat <- as.matrix(unclass(fqs))
cn <- colnames(mat)
Heatmap(mat,
    col = rev(brewer.pal(11, "RdGy")[-6]),
    name = "Frequency",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_column_names = FALSE, 
    row_names_side = "left",
    row_title = "cluster_id",
    column_title = "cl_FullLineage",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(round(mat[j, i] * 100, 1), x = x, y = y, 
            gp = gpar(col = "white", fontsize = 10)),
    bottom_annotation = HeatmapAnnotation(
      text = anno_text(cn, rot = 80, just = "right")))

n_cells_lineage <- table(sce$cl_FullLineage, sce$cluster_id)
fqs <- prop.table(n_cells_lineage, margin = 2)
mat <- as.matrix(unclass(fqs))
Heatmap(mat,
    col = rev(brewer.pal(11, "RdGy")[-6]),
    name = "Frequency",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cl_FullLineage",
    row_names_rot = 10,
    column_title = "cluster_id",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(round(mat[j, i] * 100, 1), x = x, y = y, 
            gp = gpar(col = "white", fontsize = 10)))
```



## Evaluation of cluster before and after integration

We evaluate if cells which were together in a cluster before the integration of the organoid cells are still in the same cluster after integration.

```{r cluster-before-after-organoid-integration, fig.width = 7, fig.height = 6}
## Load the Seurat object from our NSC analysis
so_before <- readRDS(file.path("output", "so_04_clustering.rds"))
so_before <- SetIdent(so_before, value = "integrated_snn_res.0.4")
so_before@meta.data$cluster_id <- Idents(so_before)
table(so_before@meta.data$cluster_id)

## subset to our cells
cs <- which(so@meta.data$integration_group %in% c("P22", "D52", "D96"))
sub <- subset(so, cells = cs)
table(sub@meta.data$cluster_id)
## join the cluster_ids from both clustering runs
before <- data.frame(cell = colnames(so_before), 
                     cluster_before = so_before@meta.data[,c("cluster_id")])
after <- data.frame(cell = colnames(sub), 
                    cluster_after = sub@meta.data[,c("cluster_id")])
clusters <- before %>% full_join(after)

## check if cells from the same cluster are still in the same cluster
(n_clusters <- table(clusters$cluster_after, clusters$cluster_before))
fqs <- prop.table(n_clusters, margin = 2)
mat <- as.matrix(unclass(fqs))
Heatmap(mat,
    col = rev(brewer.pal(11, "RdGy")[-6]),
    name = "Frequency",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "clusters after organoid integration",
    column_title = "clusters before organoid integration",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(round(mat[j, i] * 100, 2), x = x, y = y, 
            gp = gpar(col = "white", fontsize = 8)))

## add the old cluster identities to the Seurat object
so@meta.data$cluster_id_before <- so_before@meta.data$cluster_id[
  match(colnames(so), colnames(so_before))]
```


# Dimension reduction plots {.tabset} 

```{r integration-dim-red, results = "asis", fig.width = 12}
## merge the lineage labels of identical cell types
so$cl_FullLineage_merged <- as.factor(so$cl_FullLineage)
levels(so$cl_FullLineage_merged) <- c("choroid plexus/mesenchymal-like cells", 
               "cortical neurons", "cortical neurons", 
               "cycling dorsal progenitors", "cycling ventral progenitors", 
               "ectodermal/neuroectodermal-like cells", 
               "gliogenic/outer RGCs and astrocytes", 
               "IPs and early cortical neurons", "midbrain/hindbrain cells", 
               "neuroepithelial-like cells", "retina progenitors", "RGCs", 
               "RGCs early", "RGCs early", "stem cells", "stem cells", 
               "stem cells", "ventral progenitors and neurons", 
               "ventral progenitors and neurons", 
               "ventral progenitors and neurons")

cs <- sample(colnames(so), 10e3)
.plot_dr <- function(so, dr, id)
    DimPlot(so, cells = cs, group.by = id, reduction = dr, pt.size = 0.4) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
ids <- c("integration_group", "group_id", "cl_FullLineage_merged", "cluster_id", 
         "cluster_id_before")
for (id in ids) {
    cat("## ", id, "\n")
    p1 <- .plot_dr(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none")
    p2 <- .plot_dr(so, "umap", id) + theme(legend.position = "none")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.5))
    print(p)
    cat("\n\n")
}
```


# Find markers using `scran`

We identify candidate marker genes for each cluster.

```{r scran-findMarkers}
scran_markers <- findMarkers(sce, 
    groups = sce$cluster_id, block = sce$sample_id, 
    direction = "up", lfc = 2, full.stats = TRUE)
```

## Heatmap of mean marker-exprs. by cluster
We aggregate the cells to pseudobulks and plot the average expression of the candidate marker genes in each of the clusters. 
```{r scran-markers-heatmap, fig.width = 12, fig.height = 6}
gs <- lapply(scran_markers, function(u) rownames(u)[u$Top == 1])
## candidate cluster markers
lapply(gs, function(x) str_split(x, pattern = "\\.", simplify = TRUE)[,2])
sub <- sce[unique(unlist(gs)), ]
pbs <- aggregateData(sub, assay = "logcounts", by = "cluster_id", fun = "mean")
mat <- t(muscat:::.scale(assay(pbs)))
## remove the Ensembl ID from the gene names
colnames(mat) <- str_split(colnames(mat), pattern = "\\.", simplify = TRUE)[,2] 
Heatmap(mat,
    name = "scaled avg.\nexpression",
    col = viridis(10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster_id",
    rect_gp = gpar(col = "white"))
```

We write tables with the top marker genes per cluster. 
```{r}
gs2 <- lapply(scran_markers, function(u) u[u$Top %in% 1:3,])
for (i in seq_along(gs2)) {
  write.table(x = gs2[[i]] %>% as.data.frame %>% 
                dplyr::mutate(gene = rownames(gs2[[i]])) %>%
                dplyr::relocate(gene), 
              file =  file.path("output", 
                                paste0("organoid_integration_cluster", i, "_marker_genes.txt")),
              sep = "\t", quote = FALSE, row.names = FALSE)
}
```

Heatmap including marker genes of rank 2 and 3.
```{r fig.width = 14, fig.height = 6}
gs <- lapply(scran_markers, function(u) rownames(u)[u$Top %in% 1:3])
## candidate cluster markers
lapply(gs, function(x) str_split(x, pattern = "\\.", simplify = TRUE)[,2])
sub <- sce[unique(unlist(gs)), ]
pbs <- aggregateData(sub, assay = "logcounts", by = "cluster_id", fun = "mean")
mat <- t(muscat:::.scale(assay(pbs)))
## remove the Ensembl ID from the gene names
colnames(mat) <- str_split(colnames(mat), pattern = "\\.", simplify = TRUE)[,2] 
Heatmap(mat,
    name = "scaled avg.\nexpression",
    col = viridis(10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster_id",
    rect_gp = gpar(col = "white"))
```

