---
author: "Katharina Hembach"
date: "8/13/2020"
output: 
  html_document:
    toc: true,
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```

### Load packages
```{r, message = FALSE}
library(BiocParallel)
library(ggplot2)
library(dplyr)
library(cowplot)
library(ggplot2)
library(Seurat)
library(SingleCellExperiment)
library(future)
```


```{r future-setup}
# increase future's maximum allowed size of exported globals
# the default is 2GB
options(future.globals.maxSize = 11000 * 1024 ^ 2)

# change the current plan to access parallelization
plan("multiprocess", workers = 20)
```


## Load data & convert to SCE

```{r load-data}
so_nsc <- readRDS(file.path("output", "so_04_clustering.rds"))
DefaultAssay(so_nsc) <- "RNA"
```

```{r load-organoid-data}
so_org <- readRDS(file.path("output", "so_organoid-01-clustering.rds"))
DefaultAssay(so_org) <- "RNA"
```


## Clustering without integration

We merge the normalized Seurat objects of our NSCs and neural cultures and the human organoid data from Kanton et al. 2019. Then we run PCA, tSNE, UMAP and clustering. We want to know if there are big differences between the two datasets that make an integration necessary. 
```{r merge-so}
so <- merge(so_nsc, y = so_org, add.cell.ids = c("NSC", "organoid"), 
            project = "neural_cultures", merge.data = TRUE)
## variable with sample ID from our data and cell line from organoids
so$merged_sample <- ifelse(!is.na(so$Line), so$Line, so$sample_id)
## variable with group ID from our data and Stage from organoids
so$merged_group <-  ifelse(!is.na(so$group_id), so$group_id, so$Stage)
## variable with number of UMIs and fraction of MT UMIs
so$merged_nUMIs <- ifelse(!is.na(so$sum), so$sum, so$nUMI)
so$merged_fractionMt <- ifelse(!is.na(so$subsets_Mt_fraction), 
                              so$subsets_Mt_fraction, so$PercentMito)
so$merged_nGenes <- ifelse(!is.na(so$detected), so$detected, so$nGene)
```

```{r no-integration-scaling}
so <- FindVariableFeatures(so, nfeatures = 2000, 
                           selection.method = "vst", verbose = FALSE)
so <- ScaleData(so, verbose = FALSE, 
                vars.to.regress = c("merged_nUMIs", "merged_fractionMt"))
```

```{r no-integration-DR}
so <- RunPCA(so, npcs = 30, verbose = FALSE)
so <- RunTSNE(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, do.fast = TRUE, verbose = FALSE)
so <- RunUMAP(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, verbose = FALSE)

DimPlot(so, reduction = "pca", group.by = "merged_sample")

so <- FindNeighbors(so, reduction = "pca", dims = seq_len(20), verbose = FALSE)
so <- FindClusters(so, resolution = 0.4, random.seed = 1, verbose = FALSE)
```

## Dimension reduction plots {.tabset} 

We plot the dimension reduction (DR) and color by same/cell line, group/Stage, organoid cluster labels, cluster ID

```{r no-integration-dim-red, results = "asis", fig.width = 12}
# set cluster IDs to resolution 0.4 clustering
so <- SetIdent(so, value = "integrated_snn_res.0.4")
so@meta.data$cluster_id <- Idents(so)

cs <- sample(colnames(so), 10e3)
.plot_dr <- function(so, dr, id)
    DimPlot(so, cells = cs, group.by = id, reduction = dr, pt.size = 0.4) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
ids <- c("merged_sample", "merged_group","cl_FullLineage", "ident")
for (id in ids) {
    cat("### ", id, "\n")
    p1 <- .plot_dr(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none")
    p2 <- .plot_dr(so, "umap", id) + theme(legend.position = "none")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.5))
    print(p)
    cat("\n\n")
}
```


## QC on DR plots {.tabset}

```{r no-integration-DR-QC, results = "asis", fig.width = 12}
.plot_features <- function(so, dr, id) {
    FeaturePlot(so, cells = cs, features = id, reduction = dr, pt.size = 0.4, 
                cols = c("grey", "blue")) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
}
ids <- c("merged_nUMIs", "merged_nGenes", "merged_fractionMt")
for (id in ids) {
    cat("### ", id, "\n")
    p1 <- .plot_features(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none") + ggtitle("tSNE")
    p2 <- .plot_features(so, "umap", id) + theme(legend.position = "none") + 
      ggtitle("UMAP")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```


### Save Seurat object to RDS

```{r save-so-merged}
saveRDS(so, file.path("output", "so_merged_organoid-02-integration.rds"))
```


## Integration

We integrate the two datasets because they are completely separated in the DR plots (see above).

```{r integration-normalisation, warning = FALSE}
## make sure column names match between datasets:
## we will use sample_id, group_id, nUMI, fractionMt, nGene
so_nsc$nUMI <- so_nsc$sum
so_nsc$fractionMt <- so_nsc$subsets_Mt_fraction
so_nsc$nGene <- so_nsc$detected
so_nsc$sum <- NULL
so_nsc$subsets_Mt_fraction <- NULL
so_nsc$detected <- NULL
so_org$sample_id <- so_org$Line
so_org$group_id <- so_org$Stage
so_org$fractionMt <- so_org$PercentMito
so_org$Line <- NULL
so_org$Stage <- NULL
so_org$PercentMito <- NULL

# split our cells by sample
cells_by_sample <- split(colnames(so_nsc), so_nsc$sample_id)
so_nsc <- lapply(cells_by_sample, function(i) subset(so_nsc, cells = i))
# split organoid cells by cell Line
cells_by_sample <- split(colnames(so_org), so_org$sample_id)
so_org <- lapply(cells_by_sample, function(i) subset(so_org, cells = i))
## we combine the two lists
so <- c(so_nsc, so_org)

## Identify the top 2000 genes with high cell-to-cell variation
so <- lapply(so, FindVariableFeatures, nfeatures = 2000, 
    selection.method = "vst", verbose = FALSE)
```

```{r integration, warning = FALSE}
## find anchors & integrate
as <- FindIntegrationAnchors(so, verbose = FALSE)
so <- IntegrateData(anchorset = as, dims = seq_len(30), verbose = FALSE)
```

```{r integration-scale-data, warning = FALSE}
DefaultAssay(so) <- "integrated"
## We scale the data
so <- ScaleData(so, verbose = FALSE,
                vars.to.regress = c("nUMI", "fractionMt"))
```

```{r integration-DR, warning = FALSE}
so <- RunPCA(so, npcs = 30, verbose = FALSE)
so <- RunTSNE(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, do.fast = TRUE, verbose = FALSE)
so <- RunUMAP(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, verbose = FALSE)

## PCA plot 
DimPlot(so, reduction = "pca", group.by = "sample_id")

# elbow plot with the ranking of PCs based on the % of variance explained
ElbowPlot(so, ndims = 30)
```

```{r integration-clustering, warning = FALSE}
so <- FindNeighbors(so, reduction = "pca", dims = seq_len(20), verbose = FALSE)
so <- FindClusters(so, resolution = 0.4, random.seed = 1, verbose = FALSE)
```


## Dimension reduction plots {.tabset} 

We plot the dimension reduction (DR) and color by same/cell line, group/Stage, organoid cluster labels, cluster ID

```{r integration-dim-red, results = "asis", fig.width = 12}
# set cluster IDs to resolution 0.4 clustering
so <- SetIdent(so, value = "integrated_snn_res.0.4")
so@meta.data$cluster_id <- Idents(so)

cs <- sample(colnames(so), 10e3)
.plot_dr <- function(so, dr, id)
    DimPlot(so, cells = cs, group.by = id, reduction = dr, pt.size = 0.4) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
ids <- c("sample_id", "group_id", "cl_FullLineage", "ident")
for (id in ids) {
    cat("### ", id, "\n")
    p1 <- .plot_dr(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none")
    p2 <- .plot_dr(so, "umap", id) + theme(legend.position = "none")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.5))
    print(p)
    cat("\n\n")
}
```


## QC on DR plots {.tabset}

```{r integration-DR-QC, results = "asis", fig.width = 12}
.plot_features <- function(so, dr, id) {
    FeaturePlot(so, cells = cs, features = id, reduction = dr, pt.size = 0.4, 
                cols = c("grey", "blue")) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
}
ids <- c("nUMI", "nGene", "fractionMt")
for (id in ids) {
    cat("### ", id, "\n")
    p1 <- .plot_features(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none") + ggtitle("tSNE")
    p2 <- .plot_features(so, "umap", id) + theme(legend.position = "none") + 
      ggtitle("UMAP")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```

### Save Seurat object to RDS

```{r save-so}
saveRDS(so, file.path("output", "so_integrated_organoid-02-integration.rds"))
```

