---
title: "Clustering of TDP-HA experiment and the D96 samples"
author: "Katharina Hembach"
date: "28/01/2021"
output: 
  html_document:
    toc: true,
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```

### Load packages
```{r, message = FALSE}
library(BiocParallel)
library(ggplot2)
library(dplyr)
library(cowplot)
library(ggplot2)
library(Seurat)
library(SingleCellExperiment)
library(future)
```

## Load data

```{r read-data}
## Seurat objects with normalized data
so_tdp <- readRDS(file.path("output", "so_TDP_05_plasmid_expression.rds"))
so_tdp$group_id <- "TDP"
# so_timeline <- readRDS(file.path("output", "so_06-clustering_all_timepoints.rds"))

so_d96 <- readRDS(file.path("output", "so_04_clustering.rds"))
## select only the D96 cells
so_d96 <- subset(so_d96, subset = group_id == "D96")
```
We merge the samples from the two data sets into a Seurat object.
```{r merge-so}
## merge the two Seurat objects
so <- merge(so_tdp, y = so_d96, add.cell.ids = c("tdp_ha", "D96"), 
            project = "neural_cultures", merge.data = TRUE)
so$group_id <- factor(so$group_id, levels = c("D96", "TDP"))
```


## Variable features
```{r variable-features-scaling}
so <- FindVariableFeatures(so, nfeatures = 2000, 
    selection.method = "vst", verbose = FALSE)
so <- ScaleData(so, verbose = FALSE, vars.to.regress = c("sum", 
                                                         "subsets_Mt_percent"))
```


## Dimension reduction 

We perform dimension reduction with t-SNE and UMAP based on PCA results.
```{r dimension-reduction, warning = FALSE}
so <- RunPCA(so, npcs = 30, verbose = FALSE)
so <- RunTSNE(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, do.fast = TRUE, verbose = FALSE)
so <- RunUMAP(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, verbose = FALSE)
```

### Plot PCA results

```{r, fig.width = 12, fig.height = 8}
# top genes that are associated with the first two PCs
VizDimLoadings(so, dims = 1:2, reduction = "pca")
```

```{r, fig.width = 10, fig.height = 8}
## PCA plot 
DimPlot(so, reduction = "pca", group.by = "sample_id")
```

```{r}
# elbow plot with the ranking of PCs based on the % of variance explained
ElbowPlot(so, ndims = 30)
```


## Clustering

We cluster the cells using the reduced PCA dimensions.

```{r clustering, warning = FALSE}
so <- FindNeighbors(so, reduction = "pca", dims = seq_len(20), verbose = FALSE)
for (res in c(0.2, 0.4, 0.8, 1))
    so <- FindClusters(so, resolution = res, random.seed = 1, verbose = FALSE)
```


## Dimension reduction plots 

We plot the dimension reduction (DR) and color by sample, group and cluster ID
```{r dr-plots, fig.width = 10, fig.height = 9, warning = FALSE}
thm <- theme(aspect.ratio = 1, legend.position = "none")
ps <- lapply(c("sample_id", "group_id", "ident"), function(u) {
    p1 <- DimPlot(so, reduction = "tsne", group.by = u) + thm
    p2 <- DimPlot(so, reduction = "umap", group.by = u)
    lgd <- get_legend(p2)
    p2 <- p2 + thm
    list(p1, p2, lgd)
    plot_grid(p1, p2, lgd, nrow = 1,
        rel_widths = c(1, 1, 0.5))
})
plot_grid(plotlist = ps, ncol = 1)
```


## QC on DR plots {.tabset}

```{r DR-QC, results = "asis", fig.width = 12}
cs <- sample(colnames(so), 1e4) ## subsample cells
.plot_features <- function(so, dr, id) {
    FeaturePlot(so, cells = cs, features = id, reduction = dr, pt.size = 0.4, 
                cols = c("grey", "blue")) +
        guides(col = guide_colourbar()) +
        theme_void() + theme(aspect.ratio = 1)
}
ids <- c("sum", "detected", "subsets_Mt_percent", "ENSG00000120948.TARDBP", 
         "ENSG00000120948.TARDBP-alevin", "TDP43-HA")
for (id in ids) {
    cat("### ", id, "\n")
    p1 <- .plot_features(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none") + ggtitle("tSNE")
    p2 <- .plot_features(so, "umap", id) + theme(legend.position = "none") + 
      ggtitle("UMAP")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```

## Save Seurat object to RDS

```{r}
saveRDS(so, file.path("output", "so_08-00_clustering_HA_D96.rds"))
```
