---
title: "Cluster 12 analysis"
author: "Katharina Hembach"
date: "11/19/2020"
output: 
  html_document:
    toc: true,
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```

### Load packages

```{r load-libs, message = FALSE, warning = FALSE}
library(ComplexHeatmap)
library(cowplot)
library(ggplot2)
library(dplyr)
library(muscat)
library(purrr)
library(RColorBrewer)
library(viridis)
library(scran)
library(Seurat)
library(SingleCellExperiment)
library(stringr)
library(RCurl)
library(BiocParallel)
```

# Load data & convert to SCE

```{r load-data}
so <- readRDS(file.path("output", "so_TDP-06-cluster-analysis.rds"))
sce <- as.SingleCellExperiment(so, assay = "RNA")
colData(sce) <- as.data.frame(colData(sce)) %>% 
    mutate_if(is.character, as.factor) %>% 
    DataFrame(row.names = colnames(sce))
so <- SetIdent(so, value = "RNA_snn_res.0.4")
so@meta.data$cluster_id <- Idents(so)
sce$cluster_id <- Idents(so)
```


# Find markers specific for cluster 12

## Markers 

We want to find markers that are specific for cluster 12 (TDP-HA positive cells).
We are interested in markers that separate cluster 12 from ALL other clusters or only a subset of clusters.
Additionally, we compare cluster 12 to only the neuronal clusters.

We identify candidate marker genes for each cluster that enable a separation of that group from any other group. The null hypothesis is that the log FC between a cluster and the compared cluster is 2.

We first subset the data to the neuronal clusters.
```{r subset-neuronal-clusters}
neuronal_clusters <- c(0, 2:5, 7:11, 12)
sce_sub <- sce[,sce$cluster_id %in% neuronal_clusters]
dim(sce)
dim(sce_sub)
```


```{r scran-findMarkers, warning = FALSE}
## marker for any other cluster
## UP
any_up <- findMarkers(sce_sub, groups = sce_sub$cluster_id, direction = "up", 
                      lfc = 2, pval.type="any")
## top 10 markers
any_up[["12"]][1:10,1:4]
## DOWN
any_down <- findMarkers(sce_sub, groups = sce_sub$cluster_id, 
                        direction = "down", 
                        lfc = 2, pval.type="any")
any_down[["12"]][1:10,1:4]
## any default parameters
any_default <- findMarkers(sce_sub, groups = sce_sub$cluster_id, 
                        pval.type="any")
any_default[["12"]][1:10,1:4]


## markers for all other clusters
## UP
all_up <- findMarkers(sce_sub, groups = sce_sub$cluster_id, 
                      direction = "up", 
                      lfc = 2, pval.type="all")
all_up[["12"]][1:10,1:3]
## DOWN
all_down <- findMarkers(sce_sub, groups = sce_sub$cluster_id, 
                        direction = "down", 
                        lfc = 2, pval.type="all")
all_down[["12"]][1:10,1:3]
## no logFC cutoff and direction
all_default <- findMarkers(sce_sub, groups = sce_sub$cluster_id, 
                           pval.type="all")
all_default[["12"]][1:10,1:3]
```

## Heatmap of mean marker-exprs. by cluster
We aggregate the cells to pseudobulks and plot the average expression of the condidate marker genes in each of the clusters. 
```{r scran-markers-heatmap-any, fig.width = 12, fig.height = 6}
plot_marker_heatmap <- function(marker, cluster, any = TRUE, top = 1, title = "") {
  u <- marker[[cluster]]
  ## including marker genes of rank 1 to 4
  if(any){
    x <- rownames(u)[u$Top %in% 1:top]
  } else{
    x <- rownames(u)[1:top]
  }

  sub <- sce_sub[unique(x), ]
  pbs <- aggregateData(sub, assay = "logcounts", by = "cluster_id", fun = "mean")
  mat <- t(muscat:::.scale(assay(pbs)))
  ## remove the Ensembl ID from the gene names
  cnames <- colnames(mat)
  colnames(mat) <- str_split(cnames, pattern = "\\.", simplify = TRUE)[,2] 
  colnames(mat)[which(colnames(mat) == "")] <- cnames[which(colnames(mat) == "")]
  
  print(Heatmap(mat,
      name = "scaled avg.\nexpression",
      col = viridis(10),
      cluster_rows = FALSE,
      cluster_columns = FALSE,
      row_names_side = "left",
      row_title = "cluster_id",
      column_title = title,
      rect_gp = gpar(col = "white")))
}

plot_marker_heatmap(any_up, "12", top = 5, title = "any UP")
plot_marker_heatmap(any_down, "12", top = 5, title = "any DOWN")
plot_marker_heatmap(any_default, "12", top = 5, title = "any default")
```

```{r scran-markers-heatmap-all, fig.width = 12, fig.height = 6}
plot_marker_heatmap(all_up, "12", any = FALSE, top = 10, title = "all UP")
plot_marker_heatmap(all_down, "12", any = FALSE, top = 10, title = "all DOWN")
plot_marker_heatmap(all_default, "12", any = FALSE, top = 10, title = "all default")
```

### Dot plot visualization of marker expression

From Seurat github:
"The DotPlot shows scaled values (which can be both positive and negative). This is because we want to be able to visualize both highly and lowly expressed genes on the same color scale."

```{r dot-plot, fig.width = 8, fig.height = 4}
fs <- list(up = c("NPTX2", "FGF18", "TDP43-HA", "PCED1A", "MEF2A", "DYNC2H1", 
                   "APOE", "GADD45A", "BCAM", "DDIT3"),
           down = c("VGF", "SCG2", "GAP43", "C4orf48", "HINT1", "LY6H", 
                     "TUBA1A", "TMSB4X", "TUBB2B", "STMN2"))

features <- all_up[["12"]][1:10,] %>% rownames
DotPlot(so, assay = "RNA", features = features, idents = neuronal_clusters, 
        scale = TRUE, scale.min = 0, scale.max = 100, dot.scale = 7) + 
  RotatedAxis() + scale_color_viridis() + 
  theme(axis.text.x = element_text(angle=45)) + ylab("neuronal cluster ID") + 
  scale_x_discrete(name = "gene", breaks = features, labels=fs[["up"]])

features <- all_down[["12"]][1:10,] %>% rownames
DotPlot(so, assay = "RNA", features = features, idents = neuronal_clusters, 
        scale = TRUE, scale.min = 0, scale.max = 100, dot.scale = 7) + 
  RotatedAxis() + scale_color_viridis() + 
  theme(axis.text.x = element_text(angle=45)) + ylab("neuronal cluster ID") + 
  scale_x_discrete(name = "gene",  breaks = features, labels=fs[["down"]])
```

## UNC13A expression in neuronal clusters

```{r UNC13A-expression, fig.width = 3, fig.height = 4}
feature <- grep(pattern = paste0("\\.UNC13A$"), rownames(so), value = TRUE)

pdf(width = 4, height = 4)
DotPlot(so, assay = "RNA", features = feature, idents = neuronal_clusters, 
        scale = TRUE, scale.min = 0, scale.max = 100, dot.scale = 7) + 
  RotatedAxis() + scale_color_viridis() + 
  theme(axis.text.x = element_text(angle=45)) + ylab("neuronal cluster ID") + 
  scale_x_discrete(name = "gene", breaks = feature, labels = "UNC13A")
dev.off()
```

