---
title: "Quality filtering"
author: "Katharina Hembach"
date: "10/7/2020"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE)
```

### Load packages
```{r, message = FALSE}
library(scater)
library(LSD)
library(dplyr)
library(edgeR)
library(ggrepel)
```

## Load data

```{r}
sce <- readRDS(file.path("output", "sce_TDP_02_quality_control.rds"))
```

## Identification of outlier cells

Based on the QC metrics, we now identify outlier cells:

```{r outlier-identification}
cols <- c("sum", "detected", "subsets_Mt_percent")
log <- c(TRUE, TRUE, FALSE)
type <- c("both", "both", "higher")

drop_cols <- paste0(cols, "_drop")
for (i in seq_along(cols))
    colData(sce)[[drop_cols[i]]] <- isOutlier(sce[[cols[i]]], 
        nmads = 3, type = type[i], log = log[i], batch = sce$sample_id)

# Overlap of outlier cells from two metrics
sapply(drop_cols, function(i) 
    sapply(drop_cols, function(j)
        sum(sce[[i]] & sce[[j]])))

colData(sce)$discard <- rowSums(data.frame(colData(sce)[,drop_cols])) > 0
table(colData(sce)$discard)

## Plot the metrics and highlight the discarded cells
plotColData(sce, x = "sample_id", y = "sum", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "detected", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "subsets_Mt_percent", 
            colour_by = "discard")
```


We decided to additionally filter the cells in the TDP experiment samples. We use the same cutoffs as for the 96 days old neural cultures from the first experiment. We also remove the cell population with low number of UMIs and detected genes from the old neural cultures (223 days).

```{r manual-filtering}
## filter the cells with less than 5000 UMIs in the TDP experiment samples
tdp_samples <- c("TDP2wON", "TDP4wOFF", "TDP4wONa", "TDP4wONb")
colData(sce)$manual_discard_sum <- colData(sce)$sum < 5000 & 
  colData(sce)$sample_id %in% tdp_samples
## filter the cells with less than 2500 detected genes
colData(sce)$manual_discard_detected <- colData(sce)$detected < 2500 & 
  colData(sce)$sample_id %in% tdp_samples

## day 223
colData(sce)$manual_discard_sum <- colData(sce)$manual_discard_sum | 
  colData(sce)$sum < 2000 & 
  colData(sce)$sample_id %in% c("NC223a", "NC223b")
colData(sce)$manual_discard_detected <- colData(sce)$manual_discard_detected |
  colData(sce)$detected < 1500 & 
  colData(sce)$sample_id %in% c("NC223a", "NC223b")

## highlight all manually discarded cells
colData(sce)$manual_discard <- colData(sce)$manual_discard_sum |
                                   colData(sce)$manual_discard_detected
plotColData(sce, x = "sample_id", y = "sum", colour_by = "manual_discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "detected", colour_by = "manual_discard") + 
  scale_y_log10()

## highlight all discarded cells
colData(sce)$discard <- colData(sce)$manual_discard |
                                   colData(sce)$discard
plotColData(sce, x = "sample_id", y = "detected", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "sum", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "subsets_Mt_percent", 
            colour_by = "discard")
```


Plot the library size against the number of detected genes before and after filtering.
```{r, fig.width=12, fig.height = 10}
cd <- colData(sce)
layout(matrix(1:12, nrow = 3, byrow = TRUE))

for (i in levels(sce$sample_id)) {
  tmp <- cd[cd$sample_id == i,]
  heatscatter(tmp$sum, tmp$detected, log = "xy", 
              main = paste0(i, "-unfiltered"), xlab = "total counts", 
              ylab = "detected genes")
  heatscatter(tmp$sum[!tmp$discard], tmp$detected[!tmp$discard], 
              log = "xy", main = paste0(i, "-filtered"), xlab = "total counts", 
              ylab = "detected genes")    
}
```


## Removal of outlier cells
We remove the outlier cells and filter the genes:

```{r discard-outliers}
## summary of the kept cells
nr <- table(cd$sample_id)
nr_fil <- table(cd$sample_id[!cd$discard])
print(rbind(
    unfiltered = nr, filtered = nr_fil, 
    "%" = round(nr_fil / nr * 100, digits = 0)))

## discard the outlier cells
dim(sce)
sce <- sce[,!cd$discard]
dim(sce)

## we filter genes and require > 1 count in at least 20 cells
sce_filtered <- sce[rowSums(counts(sce) > 1) >= 20, ]
dim(sce_filtered)
```


## Save data to RDS

```{r}
saveRDS(sce_filtered, file.path("output", "sce_TDP_03_filtering.rds"))
saveRDS(sce, file.path("output", "sce_TDP_03_filtering_all_genes.rds"))
```

