---
title: "TDP-43-HA expression in unfiltered cells"
author: "Katharina Hembach"
date: "10/15/2020"
output: 
  html_document:
    toc: true,
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```


### Load packages

```{r, message = FALSE}
library(cowplot)
library(ggplot2)
library(Seurat)
library(SingleCellExperiment)
library(tximport)
library(scater)
library(LSD)
library(dplyr)
library(ggridges)
```

## Load data

We combine the quantification of the plasmid transcript and the endogenous TDP-43 with the CellRanger count matrix.

```{r read-sce}
sce <- readRDS(file.path("output", "sce_TDP_02_quality_control.rds"))
## we only keep the four samples of the TDP-43 experiment
sample_ids <- c("TDP4wOFF", "TDP2wON", "TDP4wONa", "TDP4wONb")
sce <- sce[,colData(sce)$sample_id %in% sample_ids]
sce$sample_id <- droplevels(sce$sample_id)
```

```{r read-alevin}
samples <- c("no1_Neural_cuture_d_96_TDP-43-HA_4w_DOXoff", 
             "no2_Neural_cuture_d_96_TDP-43-HA_2w_DOXON",
             "no3_Neural_cuture_d_96_TDP-43-HA_4w_DOXONa",
             "no4_Neural_cuture_d_96_TDP-43-HA_4w_DOXONb")
txi <- matrix(NA, nrow = 2)
for (i in 1:4) {
  fi <- file.path("data", "Sep2020", "alevin_TDP43", samples[i], 
                     "alevin/quants_mat.gz")

  # import alevin quants
  a <- tximport(fi, type="alevin")$counts
  
  ## match the alevin and CellRanger cell IDs
  colnames(a) <- paste0(colnames(a), "-1.", sample_ids[i])
  txi <- cbind(txi, a)
}
txi <- txi[,colnames(txi) != ""]
rownames(txi) <- c("ENSG00000120948.TARDBP-alevin", "TDP43-HA")
```

We add the alevin counts to the CellRanger matrix.
```{r combine-counts}
## add two new rows to counts matrix and replace the counts for matching 
## barcodes with the alevin counts
counts <- rbind(counts(sce), DelayedArray(matrix(0, nrow = 2, 
                                                 ncol = ncol(counts(sce)))))
rownames(counts) <- c(rownames(sce), rownames(txi))
## match the barcodes
colnames(txi) %in% colnames(sce) %>% table
m <- match(colnames(txi), colnames(sce))
counts[rownames(txi),m[!is.na(m)]] <- txi[,which(!is.na(m))]

# adjust rowData
rd <- rbind(rowData(sce), data.frame(ensembl_id = c("ENSG00000120948", ""), 
                                     symbol = c("TARDBP_alevin", "TDP43-HA")))
rownames(rd) <- rownames(counts)

sce <- SingleCellExperiment(list(counts=counts),
                            colData = colData(sce),
                            rowData = rd)
```

## Identification of outlier cells

Based on the QC metrics, we now identify outlier cells:

```{r outlier-identification}
cols <- c("sum", "detected", "subsets_Mt_percent")
log <- c(TRUE, TRUE, FALSE)
type <- c("both", "both", "higher")

drop_cols <- paste0(cols, "_drop")
for (i in seq_along(cols))
    colData(sce)[[drop_cols[i]]] <- isOutlier(sce[[cols[i]]], 
        nmads = 3, type = type[i], log = log[i], batch = sce$sample_id)

# Overlap of outlier cells from two metrics
sapply(drop_cols, function(i) 
    sapply(drop_cols, function(j)
        sum(sce[[i]] & sce[[j]])))

colData(sce)$discard <- rowSums(data.frame(colData(sce)[,drop_cols])) > 0
table(colData(sce)$discard)

## Plot the metrics and highlight the discarded cells
plotColData(sce, x = "sample_id", y = "sum", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "detected", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "subsets_Mt_percent", 
            colour_by = "discard")
```


We decided to additionally filter the cells in the TDP experiment samples. We use the same cutoffs as for the 96 days old neural cultures from the first experiment. We also remove the cell population with low number of UMIs and detected genes from the old neural cultures (223 days).

```{r manual-filtering}
## filter the cells with less than 5000 UMIs in the TDP experiment samples
tdp_samples <- c("TDP2wON", "TDP4wOFF", "TDP4wONa", "TDP4wONb")
colData(sce)$manual_discard_sum <- colData(sce)$sum < 5000 & 
  colData(sce)$sample_id %in% tdp_samples
## filter the cells with less than 2500 detected genes
colData(sce)$manual_discard_detected <- colData(sce)$detected < 2500 & 
  colData(sce)$sample_id %in% tdp_samples

## highlight all manually discarded cells
colData(sce)$manual_discard <- colData(sce)$manual_discard_sum |
                                   colData(sce)$manual_discard_detected
plotColData(sce, x = "sample_id", y = "sum", colour_by = "manual_discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "detected", colour_by = "manual_discard") + 
  scale_y_log10()

## highlight all discarded cells
colData(sce)$discard <- colData(sce)$manual_discard |
                                   colData(sce)$discard
plotColData(sce, x = "sample_id", y = "detected", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "sum", colour_by = "discard") + 
  scale_y_log10()
plotColData(sce, x = "sample_id", y = "subsets_Mt_percent", 
            colour_by = "discard")
```


TDP-43 expression per cell.
```{r TDP-expression}
gene_ids <- c("ENSG00000120948.TARDBP", "ENSG00000120948.TARDBP-alevin", 
              "TDP43-HA")

plotExpression(sce, gene_ids,
               x = "sample_id", exprs_values = "counts", 
               colour = "discard")

plotExpression(sce, gene_ids,
               x = "discard", exprs_values = "counts", 
               colour = "sample_id")
```

### Ridge plot

Do the filtered cells express TDP-43 and TDP-43-HA?
```{r ridge-plot, fig.width = 12}
df <- colData(sce) %>% as.data.frame() %>%
  dplyr::select(sample_id, discard, detected, sum, subsets_Mt_detected, 
                discard) %>%
  dplyr::mutate(TARDBP = as.vector(counts(sce["ENSG00000120948.TARDBP"])),
                TARDBP_alevin = as.vector(counts(sce["ENSG00000120948.TARDBP-alevin"])),
                TDP43_HA = as.vector(counts(sce["TDP43-HA"]))) 

for (g in c("TARDBP", "TARDBP_alevin", "TDP43_HA")){
  cat("#### ", g, "\n")
  p <- df %>%
    ggplot(aes(x = get(g), y = sample_id, fill = discard)) +
    geom_density_ridges(panel_scaling = FALSE, show.legend = TRUE, 
                        alpha = 0.5, color = "white", scale = 0.95, 
                        rel_min_height = 0.01) +
    # facet_wrap(~group_id, nrow = 1) + 
    theme_ridges(center_axis_labels = TRUE) + 
    scale_x_continuous(expand = c(0, 0)) + 
    xlab(g) + ggtitle("all cells")
  print(p)
  
  ## number of cells with gene count > 0
  cat("cells with gene count > 0")
  print(table(df$sample_id, df[,g] > 0))
  ## retained cells with gene count > 0
  df1 <- df %>%  dplyr::filter(get(g) > 0)
  cat("discarded cells with gene count > 0")
  print(table(df1$sample_id, df1$discard))
  
   p <- df1 %>%
    ggplot(aes(x = get(g), y = sample_id, fill = discard)) +
    geom_density_ridges(panel_scaling = FALSE, show.legend = TRUE, 
                        alpha = 0.5, color = "white", scale = 0.95, 
                        rel_min_height = 0.01) +
    # facet_wrap(~group_id, nrow = 1) + 
    theme_ridges(center_axis_labels = TRUE) + 
    scale_x_continuous(expand = c(0, 0)) + 
    xlab(g) + ggtitle(paste0("cells with ", g, " count > 0"))

  print(p)
}
```

