---
title: "Clustering of TDP-HA experiment and the D96 samples"
author: "Katharina Hembach"
date: "28/01/2021"
output: 
  html_document:
    toc: true,
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```

### Load packages
```{r, message = FALSE}
library(BiocParallel)
library(ggplot2)
library(dplyr)
library(cowplot)
library(ggplot2)
library(Seurat)
library(SingleCellExperiment)
library(future)
library(ComplexHeatmap)
library(RColorBrewer)
library(viridis)
```

## Load data

```{r read-data}
## Seurat objects with normalized data
so_tdp <- readRDS(file.path("output", "so_TDP_05_plasmid_expression.rds"))
so_tdp$group_id <- "TDP"
# so_timeline <- readRDS(file.path("output", "so_06-clustering_all_timepoints.rds"))

so_d96 <- readRDS(file.path("output", "so_04_clustering.rds"))
## select only the D96 cells
so_d96 <- subset(so_d96, subset = group_id == "D96")
```
We merge the samples from the two data sets into a Seurat object.
```{r merge-so}
## merge the two Seurat objects
so <- merge(so_tdp, y = so_d96, add.cell.ids = c("tdp_ha", "D96"), 
            project = "neural_cultures", merge.data = TRUE)
so$group_id <- factor(so$group_id, levels = c("D96", "TDP"))
```


## Variable features
```{r variable-features-scaling}
so <- FindVariableFeatures(so, nfeatures = 2000, 
    selection.method = "vst", verbose = FALSE)
so <- ScaleData(so, verbose = FALSE, vars.to.regress = c("sum", 
                                                         "subsets_Mt_percent"))
```


## Dimension reduction 

We perform dimension reduction with t-SNE and UMAP based on PCA results.
```{r dimension-reduction, warning = FALSE}
so <- RunPCA(so, npcs = 30, verbose = FALSE)
so <- RunTSNE(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, do.fast = TRUE, verbose = FALSE)
so <- RunUMAP(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, verbose = FALSE)
```

### Plot PCA results

```{r, fig.width = 12, fig.height = 8}
# top genes that are associated with the first two PCs
VizDimLoadings(so, dims = 1:2, reduction = "pca")
```

```{r, fig.width = 10, fig.height = 8}
## PCA plot 
DimPlot(so, reduction = "pca", group.by = "sample_id")
```

```{r}
# elbow plot with the ranking of PCs based on the % of variance explained
ElbowPlot(so, ndims = 30)
```


## Clustering

We cluster the cells using the reduced PCA dimensions.

```{r clustering, warning = FALSE}
so <- FindNeighbors(so, reduction = "pca", dims = seq_len(20), verbose = FALSE)
for (res in c(0.2, 0.4, 0.8, 1))
    so <- FindClusters(so, resolution = res, random.seed = 1, verbose = FALSE)
```


## Dimension reduction plots 

We plot the dimension reduction (DR) and color by sample, group and cluster ID
```{r dr-plots, fig.width = 10, fig.height = 9, warning = FALSE}
thm <- theme(aspect.ratio = 1, legend.position = "none")
ps <- lapply(c("sample_id", "group_id", "ident"), function(u) {
    p1 <- DimPlot(so, reduction = "tsne", group.by = u) + thm
    p2 <- DimPlot(so, reduction = "umap", group.by = u)
    lgd <- get_legend(p2)
    p2 <- p2 + thm
    list(p1, p2, lgd)
    plot_grid(p1, p2, lgd, nrow = 1,
        rel_widths = c(1, 1, 0.5))
})
plot_grid(plotlist = ps, ncol = 1)
```


## QC on DR plots {.tabset}

```{r DR-QC, results = "asis", fig.width = 12}
cs <- sample(colnames(so), 1e4) ## subsample cells
.plot_features <- function(so, dr, id) {
    FeaturePlot(so, cells = cs, features = id, reduction = dr, pt.size = 0.4, 
                cols = c("grey", "blue")) +
        guides(col = guide_colourbar()) +
        theme_void() + theme(aspect.ratio = 1)
}
ids <- c("sum", "detected", "subsets_Mt_percent", "ENSG00000120948.TARDBP", 
         "ENSG00000120948.TARDBP-alevin", "TDP43-HA")
for (id in ids) {
    cat("### ", id, "\n")
    p1 <- .plot_features(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none") + ggtitle("tSNE")
    p2 <- .plot_features(so, "umap", id) + theme(legend.position = "none") + 
      ggtitle("UMAP")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```

# DR with cells colored by individial clustering {.tabset}

We show the cluster membership of the individual clustering in the combined UMAP.
```{r dr-individual-clusters, results = "asis", fig.width = 12}
## combined clustering so$RNA_snn_res.0.4
## individual clustering so$integrated_snn_res.0.4 (D96)
## so_tdp$RNA_snn_res.0.4 (TDP-HA)
# we lost the individual clustering of the TDP-HA samples, 
# and will add them to the so object
so$TDPHA_snn_res.0.4 <- NA
## iterate through each sample and add the corresponding cluster ids
for( s in unique(so_tdp$sample_id)){
  ind <- so$sample_id == s
  ind_tdp <- so_tdp$sample_id == s
  so$TDPHA_snn_res.0.4[ind] <- as.character(so_tdp$RNA_snn_res.0.4[ind_tdp][match(so$barcode[ind], 
                                                      so_tdp$barcode[ind_tdp])])
}
so$TDPHA_snn_res.0.4 <- factor(so$TDPHA_snn_res.0.4, 
                               levels = as.character(0:16))

so$integrated_snn_res.0.4  <- factor(so$integrated_snn_res.0.4, 
                                     levels = as.character(0:16))

cs <- sample(colnames(so), 1e4)
.plot_dr <- function(so, dr, id)
    DimPlot(so, cells = cs, group.by = id, reduction = dr, pt.size = 0.4) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)

ids <- c("group_id", "sample_id", "ident", "TDPHA_snn_res.0.4", "integrated_snn_res.0.4")
for (id in ids) {
    cat("## ", id, "\n")
    p1 <- .plot_dr(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none")
    p2 <- .plot_dr(so, "umap", id) + theme(legend.position = "none")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```

# How are the individal clusters distributed in the combined clustering?
```{r cluster-distr-combined}
so$RNA_snn_res.0.4 %>% table
so$integrated_snn_res.0.4 %>% table
so$TDPHA_snn_res.0.4 %>% table

## D96
## check if cells from the same cluster are still in the same cluster
(n_clusters <- table(so$RNA_snn_res.0.4, so$integrated_snn_res.0.4))
fqs <- prop.table(n_clusters, margin = 2)
mat <- as.matrix(unclass(fqs))
Heatmap(mat,
    col = rev(brewer.pal(11, "RdGy")[-6]),
    name = "Frequency",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "combined clusters",
    column_title = "individual clusters",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(round(mat[j, i] * 100, 2), x = x, y = y, 
            gp = gpar(col = "white", fontsize = 8)))

## TDP-HA experiment
(n_clusters <- table(so$RNA_snn_res.0.4, so$TDPHA_snn_res.0.4))
fqs <- prop.table(n_clusters, margin = 2)
mat <- as.matrix(unclass(fqs))
Heatmap(mat,
    col = rev(brewer.pal(11, "RdGy")[-6]),
    name = "Frequency",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "combined clusters",
    column_title = "individual clusters",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(round(mat[j, i] * 100, 2), x = x, y = y, 
            gp = gpar(col = "white", fontsize = 8)))
```


# Save Seurat object to RDS

```{r}
saveRDS(so, file.path("output", "so_08-00_clustering_HA_D96.rds"))
```

