---
title: "Gene expression changes in TDP-HA cultures"
author: "Katharina Hembach"
date: "2/25/2021"
output: 
  html_document:
    toc: true,
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```

### Load packages

```{r load-libs, message = FALSE, warning = FALSE}
library(Seurat)
library(SingleCellExperiment)
library(dplyr)
library(scran)
library(Seurat)
library(ComplexHeatmap)
library(cowplot)
library(ggplot2)
library(muscat)
library(purrr)
library(RColorBrewer)
library(viridis)
library(edgeR)
library(stringr)
library(UpSetR)
library(scater)
library(BiocParallel)
```

# Load data & convert to SCE

```{r load-data}
so <- readRDS(file.path("output", "so_08-00_clustering_HA_D96.rds"))
sce <- as.SingleCellExperiment(so, assay = "RNA")
colData(sce) <- as.data.frame(colData(sce)) %>% 
    mutate_if(is.character, as.factor) %>% 
    DataFrame(row.names = colnames(sce))
```


# Cluster-sample counts

```{r}
# set cluster IDs to resolution 0.4 clustering
so <- SetIdent(so, value = "RNA_snn_res.0.4")
so@meta.data$cluster_id <- Idents(so)
sce$cluster_id <- Idents(so)
(n_cells <- table(sce$cluster_id, sce$sample_id))
```

# Relative cluster-abundances
How are the samples distributed across clusters?
```{r fig.width = 6, fig.height = 5}
fqs <- prop.table(n_cells, margin = 2)
mat <- as.matrix(unclass(fqs))
Heatmap(mat,
    col = rev(magma(12))[-c(1,2)],
    name = "Frequency",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster_id",
    column_title = "sample_id",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(round(mat[j, i] * 100, 2), x = x, y = y, 
            gp = gpar(col = "white", fontsize = 8)))
```


# DR colored by cluster ID {.tabset}

```{r dim-red, results = "asis", fig.width = 12}
cs <- sample(colnames(so), 1e4)
.plot_dr <- function(so, dr, id)
    DimPlot(so, cells = cs, group.by = id, reduction = dr, pt.size = 0.4) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1)
ids <- c("group_id", "sample_id", "ident")
for (id in ids) {
    cat("## ", id, "\n")
    p1 <- .plot_dr(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none")
    p2 <- .plot_dr(so, "umap", id) + theme(legend.position = "none")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```


# DR colored by cluster ID per sample

To better see how the cells from different clusters overlap, we only plot the cells from one samples at a time.

```{r UMAP-split-by-sample, results = "asis", fig.width = 12}

.plot_dr <- function(so, dr, id, cs) {
    DimPlot(so, cells = cs, group.by = id, reduction = dr, pt.size = 0.4, cols = ) +
        guides(col = guide_legend(nrow = 11, 
            override.aes = list(size = 3, alpha = 1))) +
        theme_void() + theme(aspect.ratio = 1) +
        theme(plot.title = element_text(hjust = 0.5))}
# ids <- unique(so$sample_id)

p1 <- .plot_dr(so, "umap", "ident", colnames(so[,so$sample_id == "5NC96"])) + 
  theme(legend.position = "none") + ggtitle("5NC96")
p2 <- .plot_dr(so, "umap", "ident", colnames(so[,so$sample_id == "6NC96"])) + 
  theme(legend.position = "none") + ggtitle("6NC96")
p3 <- .plot_dr(so, "umap", "ident", colnames(so[,so$sample_id == "TDP4wOFF"])) + 
  theme(legend.position = "none") + ggtitle("TDP4wOFF")
p4 <- .plot_dr(so, "umap", "ident", colnames(so[,so$sample_id == "TDP2wON"])) + 
  theme(legend.position = "none") + ggtitle("TDP2wON")
p5 <- .plot_dr(so, "umap", "ident", colnames(so[,so$sample_id == "TDP4wONa"])) + 
  theme(legend.position = "none") + ggtitle("TDP4wONa")
p6 <- .plot_dr(so, "umap", "ident", colnames(so[,so$sample_id == "TDP4wONb"])) + 
  theme(legend.position = "none") + ggtitle("TDP4wONb")
ps <- plot_grid(plotlist = list(p1, p2, p3, p4, p5, p6), nrow = 2)
lgd <- get_legend(p5)
p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
p
```


# DE analysis

We want to test for differences in gene expression between cells from D96 samples and the TDP-HA expressing samples. We test for differences in the neuronal cells (right cell cloud in UMAP).

## Preprocessing with muscat

We filter and only keep cells from D96 and 4wON.

```{r muscat-preprocessing}
sce$id <- sce$sample_id
levels(sce$id) <- c("D96", "D96", "ON2w", "OFF4w", "ON4w", "ON4w")
(sce <- prepSCE(sce, 
    kid = "cluster_id", # subpopulation assignments
    gid = "id",  # group IDs (ctrl/stim)
    sid = "sample_id",   # sample IDs (ctrl/stim.1234)
    drop = FALSE))  # drop all other colData columns

nk <- length(kids <- levels(sce$cluster_id))
ns <- length(sids <- levels(sce$sample_id))
names(kids) <- kids; names(sids) <- sids

# nb. of cells per cluster-sample
t(table(sce$cluster_id, sce$sample_id))
```


## Compute pseudobulk counts

We sum the gene counts per cluster
```{r pseudobulk}
pb <- aggregateData(sce, assay = "counts", by = c("cluster_id", "sample_id"), 
                    fun = "sum")

# one sheet per subpopulation = cluster
assayNames(pb)
# pseudobulks for 1st cluster
t(head(assay(pb)))
```


## Pseudobulk MDS plot

```{r MDS}
## modified version of the muscat function to plot different MDS dimensions 
pbMDS_mod <- function(x, dim1 = 1, dim2 = 2) {
    # check validity of input pseudobulk-SCE
    # (cells should have been aggregated by cluster-sample)
    muscat:::.check_pbs(x, check_by = TRUE)
    
    y <- as.list(assays(x))
    y <- do.call("cbind", y)
    y <- y[, (j <- c(t(metadata(x)$n_cells)) != 0)]
    d <- DGEList(unname(y), remove.zeros = TRUE)
    d <- calcNormFactors(d)
    
    mds <- plotMDS.DGEList(d, plot = FALSE, dim.plot = c(dim1, dim2))
    nk <- length(kids <- assayNames(x))

    ss <- rep(colnames(x), nk)
    ks <- rep(kids, each = ncol(x))
    
    if (any(!j)) {
        txt <- paste(sQuote(ks[!j]), sQuote(ss[!j]), sep = "-")
        message("Removing cluster-sample instance(s) ", 
            paste(txt, collapse = ", "))
    }

    df <- data.frame(
        MDS1 = mds$x, MDS2 = mds$y, 
        cluster_id = factor(ks[j], levels = kids), 
        group_id = rep(x$group_id, nk)[j])
    
    cols <- muscat:::.cluster_colors
    if (nk > length(cols)) 
        cols <- colorRampPalette(cols)(nk)
    
    ggplot(df, aes_string(x = "MDS1", y = "MDS2", 
        col = "cluster_id", shape = "group_id")) +
        scale_color_manual(values = cols) +
        geom_point(size = 3, alpha = 0.8) + 
        guides(color = guide_legend(override.aes = list(alpha = 1))) +
        theme_bw() + theme(aspect.ratio = 1,
            axis.text = element_text(color = "black"),
            panel.grid.minor = element_blank(),
            panel.grid.major = element_line(size = 0.2, color = "lightgrey")) + 
        labs(x = paste0("MDS", dim1), y = paste0("MDS", dim2))
}

pbMDS(pb) + guides(col = guide_legend(nrow = 10)) + theme_bw(base_size = 14) 
pbMDS_mod(pb, 1, 3)  +  guides(col = guide_legend(nrow = 10)) + 
  theme_bw(base_size = 14) 
pbMDS_mod(pb, 1, 4) + guides(col = guide_legend(nrow = 10)) + 
  theme_bw(base_size = 14) 
```

# Subsetting of D96 and ON4w samples

We only keep the two replicates of the D96 and ON4w groups because ON2w and OFF4w do not have any replicates.
```{r subsetting}
sce_sub <- sce[, sce$group_id %in% c("D96", "ON4w")]
sce_sub$sample_id <- droplevels(sce_sub$sample_id)
sce_sub$id <- sce_sub$sample_id
levels(sce_sub$id) <- c("D96", "D96", "ON4w", "ON4w")
(sce_sub <- prepSCE(sce_sub, 
    kid = "cluster_id", # subpopulation assignments
    gid = "id",  # group IDs (ctrl/stim)
    sid = "sample_id",   # sample IDs (ctrl/stim.1234)
    drop = FALSE))  # drop all other colData columns

nk <- length(kids <- levels(sce_sub$cluster_id))
ns <- length(sids <- levels(sce_sub$sample_id))
names(kids) <- kids; names(sids) <- sids

# nb. of cells per cluster-sample
t(table(sce_sub$cluster_id, sce_sub$sample_id))
```


## Compute pseudobulk counts

We sum the gene counts per cluster.
```{r pseudobulk-sub}
pb <- aggregateData(sce_sub, assay = "counts", by = c("cluster_id", "sample_id"), 
                    fun = "sum")

# one sheet per subpopulation = cluster
assayNames(pb)
# pseudobulks for 1st cluster
t(head(assay(pb)))
```


## Pseudobulk MDS plot

```{r MDS-sub}
pbMDS(pb) + guides(col = guide_legend(nrow = 10)) + theme_bw(base_size = 14) 
pbMDS_mod(pb, 1, 3)  +  guides(col = guide_legend(nrow = 10)) + 
  theme_bw(base_size = 14) 
pbMDS_mod(pb, 1, 4) + guides(col = guide_legend(nrow = 10)) + 
  theme_bw(base_size = 14) 
```


## Pseudobulk differential state analysis

```{r ds-pseudobulk}
# run DS analysis
# construct design & contrast matrix
ei <- metadata(sce_sub)$experiment_info
mm <- model.matrix(~ 0 + ei$group_id)
dimnames(mm) <- list(ei$sample_id, levels(ei$group_id))
contrast <- makeContrasts("ON4w-D96", levels = mm)
mm
contrast
# run DS analysis
res <- pbDS(pb, design = mm, contrast = contrast, verbose = FALSE)

# access results table for 1st comparison
tbl <- res$table[[1]]
# one data.frame per cluster
names(tbl)
# view results for 1st cluster
k1 <- tbl[[1]]
head(format(k1[, -ncol(k1)], digits = 2))
```


## Results filtering

```{r filter-results}
# filter FDR < 5%, abs(logFC) > 1 & sort by adj. p-value
tbl_fil <- lapply(tbl, function(u) {
  u <- dplyr::filter(u, p_adj.loc < 0.01, abs(logFC) > 1)
  dplyr::arrange(u, p_adj.loc)
})

# nb. of DS genes & % of total by cluster
n_de <- vapply(tbl_fil, nrow, numeric(1))
p_de <- format(n_de / nrow(sce_sub) * 100, digits = 3)
data.frame("#DS" = n_de, "%DS" = p_de, check.names = FALSE)

# view top 3 hits in each cluster
top2 <- bind_rows(lapply(tbl_fil, slice_min, p_adj.loc, n = 3))
# top2$gene <- str_split(top2$gene, pattern = "\\.", simplify = TRUE)[,2]
# increase line width for printing of the full table
options(width = 100)
format(top2[, -ncol(top2)], digits = 2)
```


## Calculating expression frequency
We filter out genes that are not expressed in at least 10% of cells of one group.
```{r expression-freq}
frq <- calcExprFreqs(sce_sub, assay = "counts", th = 0)
# one sheet per cluster
assayNames(frq)
# expression frequencies in each
# sample & group; 1st cluster
t(head(assay(frq), 5))

# We can use the obtained frequencies to, for instance, only retain genes that 
# are expressed in an average of 10% of cells in at least 1 group:
gids <- levels(sce_sub$group_id)
frq10 <- vapply(as.list(assays(frq)), 
  function(u) apply(u[, gids] > 0.1, 1, any), 
  logical(nrow(sce_sub)))
t(head(frq10))

tbl_fil2 <- lapply(names(tbl_fil), function(k)
  dplyr::filter(tbl_fil[[k]], 
    gene %in% names(which(frq10[, k]))))
names(tbl_fil2) <- names(tbl_fil)

# nb. of DS genes & % of total by cluster
n_de <- vapply(tbl_fil2, nrow, numeric(1))
p_de <- format(n_de / nrow(sce_sub) * 100, digits = 3)
data.frame("#DS" = n_de, "%DS" = p_de, check.names = FALSE)

top2 <- bind_rows(lapply(tbl_fil2, slice_min, p_adj.loc, n = 3))
format(top2[, -ncol(top2)], digits = 2)
```


# Visualization

## Between-cluster concordance

```{r upset, fig.width=20}
de_gs_by_k <- map(tbl_fil2, "gene")
upset(fromList(de_gs_by_k), nsets = length(de_gs_by_k), nintersects = 40)
upset(fromList(de_gs_by_k))
```

## DR colored by expression
```{r plot-top-genes, fig.width = 12, fig.height = 12}
.plot_dr <- function(sce, dr, col)
  plotReducedDim(sce, dimred = dr, colour_by = col) +
    guides(fill = guide_legend(override.aes = list(alpha = 1, size = 1))) +
    theme_minimal() + theme(aspect.ratio = 1)


# downsample to max. 100 cells per cluster
cs_by_k <- split(colnames(sce_sub), sce_sub$cluster_id)
cs100 <- unlist(sapply(cs_by_k, function(u) 
  sample(u, min(length(u), 100))))

## pull the top 2 genes of each cluster
top <-  bind_rows(tbl_fil2) %>% group_by(cluster_id) %>% 
  dplyr::arrange(p_adj.loc) %>% 
  slice_min(order_by = p_adj.loc, n = 2) 
print(top, n = nrow(top))
top <- top %>% pull("gene") %>% unique

# for ea. gene in 'top8', plot t-SNE colored by its expression 
ps <- lapply(top, function(g)
  .plot_dr(sce_sub[, cs100], "UMAP", g) + 
    ggtitle(if(length(str_split(g, "\\.", simplify = TRUE)) ==2) 
      str_split(g, "\\.", simplify = TRUE)[,2] else g) + 
    theme(legend.position = "none"))

# arrange plots
plot_grid(plotlist = ps, ncol = 4, align = "vh")
```

## Cell-level visualization {.tabset}
Plot the top 6 
```{r cell-level-plot,results = "asis"}
for(i in names(tbl_fil2)) {
   cat("### ", i, "\n")
  p <- plotExpression(sce[, sce$cluster_id == i],
  features = tbl_fil2[[i]]$gene[seq_len(6)],
  x = "sample_id", colour_by = "group_id", ncol = 3) +
  guides(fill = guide_legend(override.aes = list(size = 5, alpha = 1))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p)
  cat("\n\n")
}

```

## Sample-level visualization

```{r sample-level, fig.width = 10, fig.height = 14}
pbHeatmap(sce_sub, res, top_n = 5)
```


# Include the 4wOFF but not the 2wON sample
We include the 4wOFF sample and correct for batch effects.

## Pseudobulk differential state analysis

```{r subsetting2}
sce_sub <- sce[, sce$group_id %in% c("D96", "OFF4w", "ON4w")]
sce_sub$sample_id <- droplevels(sce_sub$sample_id)
sce_sub$id <- sce_sub$sample_id
levels(sce_sub$id) <- c("no_TDP_HA", "no_TDP_HA", "no_TDP_HA", "TDP_HA", "TDP_HA")

(sce_sub <- prepSCE(sce_sub, 
    kid = "cluster_id", # subpopulation assignments
    gid = "id",  # group IDs (ctrl/stim)
    sid = "sample_id",   # sample IDs (ctrl/stim.1234)
    drop = FALSE))  # drop all other colData columns

nk <- length(kids <- levels(sce_sub$cluster_id))
ns <- length(sids <- levels(sce_sub$sample_id))
names(kids) <- kids; names(sids) <- sids

# nb. of cells per cluster-sample
t(table(sce_sub$cluster_id, sce_sub$sample_id))
```


```{r ds-pseudobulk-2}
metadata(sce_sub)$experiment_info$batch <- factor(c("TDP_exp", "TDP_exp", "TDP_exp", "WT", "WT" ))

## get pseudocounts
pb <- aggregateData(sce_sub, assay = "counts", by = c("cluster_id", "sample_id"), 
                    fun = "sum")
# tabulate number of cells
cd <- data.frame(colData(sce_sub)[, "cluster_id"])
for (i in names(cd))
    if (is.factor(cd[[i]]))
        cd[[i]] <- droplevels(cd[[i]])
ns <- table(cd)
ns <- c(unclass(ns))
metadata(pb)$n_cells <- ns
    
# construct design & contrast matrix
ei <- metadata(sce_sub)$experiment_info
mm <- model.matrix(~ batch + group_id, data = ei)
rownames(mm) <- ei$sample_id
mm
## edgeR automatically test the last colum if we don't specify a contrast, 
## i.e. TDP-HA vs. no TDP-HA
```

```{r helper-function}
.n_cells_mod <- function(x) {
    y <- metadata(x)$n_cells
    if (is.null(y)) return(NULL)
    if (length(metadata(x)$agg_pars$by) == 2)
        y <- as.matrix(data.frame(y, check.names = FALSE))
    return(as.table(y))
}
```

```{r pbDS-function}
# this is the pbDS code, fixed so it runs for our data
method <- "edgeR"
filter <- "both"
min_cells <- 10
treat <- FALSE
# muscat:::.check_pbs(pb, check_by = TRUE)
BPPARAM <- SerialParam(progressbar = TRUE)
design <- mm
coef <- NULL
contrast <- NULL

if (is.null(design)) {
    formula <- ~ group_id
    cd <- as.data.frame(colData(pb))
    design <- model.matrix(formula, cd)
    colnames(design) <- levels(pb$group_id)
    # args$design <- design
}
if (is.null(coef) & is.null(contrast)) {
    c <- colnames(design)[ncol(design)]
    contrast <- makeContrasts(contrasts = c, levels = design)
    # args$contrast <- contrast
}

# ct: type of comparison - "contrast" or "coef"
# cs: named list of 'coef's or 'contrast's
if (!is.null(contrast)) {
    coef <- NULL
    names(cs) <- cs <- colnames(contrast)
} else if (!is.null(coef)) {
    if (!is.list(coef)) 
        coef <- list(coef)
    cs <- vapply(coef, function(i)
        paste(colnames(design)[i], collapse = "-"),
        character(1))
    names(cs) <- names(coef) <- cs
}
ct <- ifelse(is.null(coef), "contrast", "coef")

if (!is.function(method)) {
    fun <- switch(method,
        "DESeq2" = muscat:::.DESeq2,
        "edgeR" = muscat:::.edgeR, 
        "limma-trend" = muscat:::.limma_trend, 
        "limma-voom" = muscat:::.limma_voom)
} else {
    fun_call <- 1
}
fun_args <- names(as.list(args(fun)))
fun_args <- fun_args[-length(fun_args)]

# for ea. cluster, run DEA
n_cells <- .n_cells_mod(pb)
names(kids) <- kids <- assayNames(pb)
res <- bplapply(
    BPPARAM = BPPARAM, 
    kids, function (k) {
    rmv <- n_cells[k, ] < min_cells
    d <- design[colnames(y <- pb[ , !rmv]), , drop = FALSE]
    if (filter %in% c("samples", "both")) {
        ls <- colSums(assay(y, k))
        ol <- isOutlier(ls, log = TRUE, type = "lower", nmads = 3)
        d <- d[colnames(y <- y[, !ol]), , drop = FALSE]
    }
    if (any(tabulate(y$group_id) < 2) 
        || qr(d)$rank == nrow(d) 
        || qr(d)$rank < ncol(d)) 
        return(NULL)
    y <- y[rowSums(assay(y, k)) != 0, ]
    if (filter %in% c("genes", "both") & max(assay(y, k)) > 100) 
        y <- y[filterByExpr(assay(y, k), d), ]
    args <- list(x = y, k = k, design = d, coef = coef, 
        contrast = contrast, ct = ct, cs = cs, treat = treat)
    args <- args[intersect(names(args), fun_args)]
    suppressWarnings(do.call(fun, args))
})

# remove empty clusters
rmv <- vapply(res, is.null, logical(1))
res <- res[!rmv]
kids <- kids[names(res)]

# reorganize & do global p-value adjustment
names(i) <- i <- c("table", "data", "fit")
res <- lapply(i, map, .x = res)
res$table <- muscat:::.p_adj_global(res$table)
# return(c(res, list(args = args)))
res2 <- c(res, list(args = args))

# access results table for 1st comparison
tbl <- res2$table[[1]]
# one data.frame per cluster
names(tbl)
# view results for 1st cluster
k1 <- tbl[[1]]
head(format(k1[, -ncol(k1)], digits = 2))
```


## Results filtering

```{r filter-results-all}
# filter FDR < 5%, abs(logFC) > 1 & sort by adj. p-value
tbl_fil <- lapply(tbl, function(u) {
  u <- dplyr::filter(u, p_adj.loc < 0.05, abs(logFC) > 0.5)
  dplyr::arrange(u, p_adj.loc)
})
## remove clusters without genes after filtering
tbl_fil <- tbl_fil[lapply(tbl_fil, nrow) > 0]

# nb. of DS genes & % of total by cluster
n_de <- vapply(tbl_fil, nrow, numeric(1))
p_de <- format(n_de / nrow(sce) * 100, digits = 3)
data.frame("#DS" = n_de, "%DS" = p_de, check.names = FALSE)

# view top 3 hits in each cluster
top2 <- bind_rows(lapply(tbl_fil, slice_min, p_adj.loc, n = 3))
# top2$gene <- str_split(top2$gene, pattern = "\\.", simplify = TRUE)[,2]
# increase line width for printing of the full table
options(width = 100)
format(top2[, -ncol(top2)], digits = 2)
```


## Calculating expression frequency
We filter out genes that are not expressed in at least 10% of cells of one group.
```{r expression-freq-all}
frq <- calcExprFreqs(sce_sub, assay = "counts", th = 0)
# one sheet per cluster
assayNames(frq)
# expression frequencies in each
# sample & group; 1st cluster
t(head(assay(frq), 5))

# We can use the obtained frequencies to, for instance, only retain genes that 
# are expressed in an average of 10% of cells in at least 1 group:
gids <- levels(sce_sub$group_id)
frq10 <- vapply(as.list(assays(frq)), 
  function(u) apply(u[, gids] > 0.1, 1, any), 
  logical(nrow(sce_sub)))
t(head(frq10))

tbl_fil2 <- lapply(names(tbl_fil), function(k)
  dplyr::filter(tbl_fil[[k]], 
    gene %in% names(which(frq10[, k]))))
names(tbl_fil2) <- names(tbl_fil)
## remove clusters without genes after filtering
tbl_fil2 <- tbl_fil2[lapply(tbl_fil2, nrow) > 0]

# nb. of DS genes & % of total by cluster
n_de <- vapply(tbl_fil2, nrow, numeric(1))
p_de <- format(n_de / nrow(sce_sub) * 100, digits = 3)
data.frame("#DS" = n_de, "%DS" = p_de, check.names = FALSE)

top2 <- bind_rows(lapply(tbl_fil2, slice_min, p_adj.loc, n = 3))
format(top2[, -ncol(top2)], digits = 2)
```


# Visualization

## Between-cluster concordance

```{r upset-all, fig.width=20, fig.height = 10}
de_gs_by_k <- map(tbl_fil2, "gene")
upset(fromList(de_gs_by_k), nsets = length(de_gs_by_k), nintersects = 40)
```

## DR colored by expression
```{r plot-top-genes-all, fig.width = 12, fig.height = 12}
# downsample to max. 100 cells per cluster
cs_by_k <- split(colnames(sce), sce$cluster_id)
cs100 <- unlist(sapply(cs_by_k, function(u) 
  sample(u, min(length(u), 100))))

## pull the top 2 genes of each cluster
top <-  bind_rows(tbl_fil2) %>% group_by(cluster_id) %>% 
  dplyr::arrange(p_adj.loc) %>% 
  slice_min(order_by = p_adj.loc, n = 2) 
print(top, n = nrow(top))
top <- top %>% pull("gene") %>% unique

# for ea. gene in 'top8', plot t-SNE colored by its expression 
ps <- lapply(top, function(g)
  .plot_dr(sce[, cs100], "UMAP", g) + 
    ggtitle(if(length(str_split(g, "\\.", simplify = TRUE)) ==2) 
      str_split(g, "\\.", simplify = TRUE)[,2] else g) + 
    theme(legend.position = "none"))

# arrange plots
plot_grid(plotlist = ps, ncol = 3, align = "vh")
```

## Cell-level visualization {.tabset}
Plot the top 6 
```{r cell-level-plot-all, results = "asis"}
for(i in names(tbl_fil2)) {
   cat("### ", i, "\n")
  p <- plotExpression(sce[, sce$cluster_id == i],
  features = tbl_fil2[[i]]$gene[if (length(tbl_fil2[[i]]$gene) < 6) {
    seq_len(length(tbl_fil2[[i]]$gene))}
    else {seq_len(6)}],
  x = "sample_id", colour_by = "group_id", ncol = 3) +
  guides(fill = guide_legend(override.aes = list(size = 5, alpha = 1))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p)
  cat("\n\n")
}
```


## Sample-level visualization

```{r sample-level-all, fig.width = 10, fig.height = 14}
pbHeatmap(sce, res2, k = c("5", "7", "10", "12", "13", "14"), top_n = 5)
# top-20 DS genes for cluster 13
pbHeatmap(sce, res2, k = "13")
```

# Save results to RDS

```{r save-rds}
saveRDS(res2, file.path("output", "res_TDP-08-01-muscat.rds"))
saveRDS(sce, file.path("output", "sce_TDP-08-01-muscat.rds"))
saveRDS(tbl_fil2, file.path("output", "tbl_TDP-08-01-muscat.rds"))
```

