---
title: "Total TDP and construct quantification"
author: "Katharina Hembach"
date: "22.07.2021"
output: 
  html_document:
    toc: true,
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```


### Load packages

```{r load-libs, message = FALSE, warning = FALSE}
library(tximport)
library(ComplexHeatmap)
library(cowplot)
library(ggplot2)
library(dplyr)
library(muscat)
library(purrr)
library(RColorBrewer)
library(viridis)
library(scran)
library(Seurat)
library(SingleCellExperiment)
library(stringr)
```

# Load data 

```{r load-data}
so <- readRDS(file.path("output", "so_TDP-06-cluster-analysis.rds"))
so <- SetIdent(so, value = "RNA_snn_res.0.4")
```


We combine the quantification of the total TDP, the construct, STMN2 and VIM with the CellRanger count matrix.
```{r read-alevin}
sample_ids <- c("TDP4wOFF", "TDP2wON", "TDP4wONa", "TDP4wONb")
samples <- c("no1_Neural_cuture_d_96_TDP-43-HA_4w_DOXoff", 
             "no2_Neural_cuture_d_96_TDP-43-HA_2w_DOXON",
             "no3_Neural_cuture_d_96_TDP-43-HA_4w_DOXONa",
             "no4_Neural_cuture_d_96_TDP-43-HA_4w_DOXONb")
txi <- matrix(NA, nrow = 4)
for (i in 1:4) {
  fi <- file.path("data", "Sep2020", "alevin_total_TDP43", samples[i], 
                     "alevin/quants_mat.gz")

  # import alevin quants
  a <- tximport(fi, type="alevin")$counts
  
  ## match the alevin and CellRanger cell IDs
  colnames(a) <- paste0(colnames(a), "-1.", sample_ids[i])
  txi <- cbind(txi, a)
}
txi <- txi[,colnames(txi) != ""]
```

We add the alevin counts to the CellRanger matrix.
```{r combine-counts}
## add two new rows to counts matrix and replace the counts for matching 
## barcodes with the alevin counts
alevin_counts <- matrix(0, nrow = 4, ncol = ncol(so))
colnames(alevin_counts) <- colnames(so)
rownames(alevin_counts) <- rownames(txi)
## match the barcodes
m <- match(colnames(txi), colnames(so))
for(i in rownames(txi)){
alevin_counts[i,m[!is.na(m)]] <- txi[i,which(!is.na(m))]
}

## add new assay with the alevin quantifications
so[["alevin"]] <- CreateAssayObject(counts = alevin_counts)

## we estimate a scaling factor for the alevin counts so they are ona similar scale than the CellRanger counts
(ratio <- (colSums(so) %>% median / colSums(so[["alevin"]]) %>% median))
(sf <- ceiling(10000/ratio))
DefaultAssay(so) <- "alevin"
## normalization with the scale factor proprtional to the difference in counts between the two assays
so <- NormalizeData(so, verbose = FALSE, scale.factor = sf, 
             normalization.method = "LogNormalize")
DefaultAssay(so) <- "RNA"
```


# DR colored by marker expression {.tabset}

```{r dim-red-expr, results = "asis", fig.width = 12, fig.height = 4}
# downsample to 5000 cells
cs <- sample(colnames(so), 5e3)
sub <- subset(so, cells = cs)

## plot the expression of the endogenous TDP-43 and TDP-HA
tdp <- c("ENSG00000120948.TARDBP", "ENSG00000120948.TARDBP-alevin", "TDP43-HA")
names(tdp) <- c("TARDBP", "TARDBP-alevin", "TDP-HA")
cat("## TDP-43\n")
ps <- lapply(seq_along(tdp), function(i) {
    if (!tdp[i] %in% rownames(sub)) return(NULL)
    FeaturePlot(sub, features = tdp[i], reduction = "umap", pt.size = 0.4,
                slot = "data") +
        theme(aspect.ratio = 1, legend.position = "none") +
        ggtitle(names(tdp)[i]) + theme_void() + theme(aspect.ratio = 1)
})
# arrange plots in grid
ps <- ps[!vapply(ps, is.null, logical(1))]
p <- plot_grid(plotlist = ps, ncol = 4, label_size = 10)
print(p)
cat("\n\n")

DefaultAssay(sub) <- "alevin"
## total TDP and control genes
tdp <- c("total-TDP43", "construct")
names(tdp) <- c("total-TDP43", "construct")
cat("## total TDP-43\n")
ps <- lapply(seq_along(tdp), function(i) {
    if (!tdp[i] %in% rownames(sub)) return(NULL)
    FeaturePlot(sub, features = tdp[i], reduction = "umap", pt.size = 0.4,
                slot = "data") +
        theme(aspect.ratio = 1, legend.position = "none") +
        ggtitle(names(tdp)[i]) + theme_void() + theme(aspect.ratio = 1)
})
# arrange plots in grid
ps <- ps[!vapply(ps, is.null, logical(1))]
p <- plot_grid(plotlist = ps, ncol = 4, label_size = 10)
print(p)
cat("\n\n")

## control genes STMN2 and VIM
g <- c("STMN2-alevin", "VIM-alevin")
names(g) <- c("STMN2-alevin", "VIM-alevin")
cat("## control genes alevin\n")
ps <- lapply(seq_along(g), function(i) {
    if (!g[i] %in% rownames(sub)) return(NULL)
    FeaturePlot(sub, features = g[i], reduction = "umap", pt.size = 0.4,
                slot = "data") +
        theme(aspect.ratio = 1, legend.position = "none") +
        ggtitle(names(g)[i]) + theme_void() + theme(aspect.ratio = 1)
})
# arrange plots in grid
ps <- ps[!vapply(ps, is.null, logical(1))]
p <- plot_grid(plotlist = ps, ncol = 4, label_size = 10)
print(p)
cat("\n\n")

DefaultAssay(sub) <- "RNA"
g <- c("ENSG00000104435.STMN2", "ENSG00000026025.VIM")
names(g) <- c("STMN2", "VIM")
cat("## control genes CellRanger\n")
ps <- lapply(seq_along(g), function(i) {
    if (!g[i] %in% rownames(sub)) return(NULL)
    FeaturePlot(sub, features = g[i], reduction = "umap", pt.size = 0.4,
                slot = "data") +
        theme(aspect.ratio = 1, legend.position = "none") +
        ggtitle(names(g)[i]) + theme_void() + theme(aspect.ratio = 1)
})
# arrange plots in grid
ps <- ps[!vapply(ps, is.null, logical(1))]
p <- plot_grid(plotlist = ps, ncol = 4, label_size = 10)
print(p)
cat("\n\n")
```


# Heatmap with TDP, construct and control genes

```{r prepare-sce}
## prepare sce object with all genes in one count matrix!
merged_counts <- rbind(so@assays$RNA@counts, so@assays$alevin@counts)
sce <- SingleCellExperiment(list(counts=merged_counts, 
                                 logcounts = rbind(so@assays$RNA@data, 
                                                   so@assays$alevin@data)))
sce$cluster_id <- Idents(so)
sce$sample_id <- so$sample_id
```

Apart from the usual marker genes, we also want to analyse the expression of Casein Kinase 1 Epsilon (CSNK1E).

```{r}
fs <- list(TDP = c("ENSG00000120948.TARDBP", "ENSG00000120948.TARDBP-alevin", 
                   "TDP43-HA", "total-TDP43", "construct"),
           control = c("ENSG00000104435.STMN2", "STMN2-alevin", 
                       "ENSG00000026025.VIM", "VIM-alevin"))
fs <- lapply(fs, function(x) unlist(x[lengths(x) !=0]) )
gs <- gsub(".*\\.", "", unlist(fs))
ns <- vapply(fs, length, numeric(1))
ks <- rep.int(names(fs), ns)
labs <- lapply(fs, function(x) gsub(".*\\.", "",x))
```

```{r known-markers-heatmap, fig.width = 10, fig.height = 18}
# split cells by cluster
cs_by_k <- split(colnames(sce), sce$cluster_id)
# compute cluster-marker means
ms_by_cluster <- lapply(fs, function(gs) vapply(cs_by_k, function(i)
        Matrix::rowMeans(logcounts(sce)[gs, i, drop = FALSE]), 
        numeric(length(gs))))

# prep. for plotting & scale b/w 0 and 1
mat <- do.call("rbind", ms_by_cluster)
mat <- muscat:::.scale(mat)
rownames(mat) <- gs
cols <- muscat:::.cluster_colors[seq_along(fs)]
cols <- setNames(cols, names(fs))
row_anno <- rowAnnotation(
    df = data.frame(label = factor(ks, levels = names(fs))),
    col = list(label = cols), gp = gpar(col = "white"))
# percentage of cells from each of the samples per cluster
(n_cells <- table(sce$cluster_id, sce$sample_id))
sample_props <- prop.table(n_cells, margin = 1)
col_mat <- as.matrix(unclass(sample_props))
sample_cols <- c("#882255", "#11588A",  "#117733", "#44AA99")
sample_cols <- setNames(sample_cols, colnames(col_mat))
col_anno <- HeatmapAnnotation(
    perc_sample = anno_barplot(col_mat, gp = gpar(fill = sample_cols), 
                               height = unit(2, "cm"),
                               border = FALSE),
    annotation_label = "fraction of sample\nin cluster",
    gap = unit(10, "points"))
col_lgd <- Legend(labels = names(sample_cols),
       title = "sample",
       legend_gp = gpar(fill = sample_cols))

hm <- Heatmap(mat,
    name = "scaled avg.\nexpression",
    col = viridis(10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    column_title = "cluster_id",
    column_title_side = "bottom",
    column_names_side = "bottom",
    column_names_rot = 0, 
    column_names_centered = TRUE,
    rect_gp = gpar(col = "white"),
    left_annotation = row_anno,
    top_annotation = col_anno)
draw(hm, annotation_legend_list = list(col_lgd))
```



# Adding the new alevin quantifications to all other 
We don't separate the new alevin quantifications but add them to the count matrix with the CellRanger quantifications.
```{r merge_counts-alevin}
## new seurat object with the merged counts
so_merged <- CreateSeuratObject(
    counts = merged_counts,
    meta.data = so[[]], ## so@meta.data
    project = "TDP_experiment")
```

## Normalization
```{r normalization-merged-counts, warning = FALSE}
# split by sample
cells_by_sample <- split(colnames(so_merged), so_merged$sample_id)
so_merged <- lapply(cells_by_sample, function(i) subset(so_merged, cells = i))

## log normalize the data using a scaling factor of 10000
so_merged <- lapply(so_merged, NormalizeData, verbose = FALSE, scale.factor = 10000, 
             normalization.method = "LogNormalize")
```


We merge the normalized and data of the six samples into a combined Seurat object and compute variable features.
```{r merge-samples}
## merge the individial Seurat objects and conserve the normalized and scaled data
so_merged <- merge(so_merged[[1]], y = so_merged[2:length(so_merged)], project = "TDP_experiment", 
            merge.data = TRUE)
## use previously computed dimension reduction
so_merged@reductions <- so@reductions
so_merged$RNA_snn_res.0.4 <- factor(so_merged$RNA_snn_res.0.4, levels = 0:16)
so_merged <- SetIdent(so_merged, value = "RNA_snn_res.0.4")
```


# DR colored by marker expression {.tabset}

```{r dim-red-expr-merged, results = "asis", fig.width = 12, fig.height = 4}
# downsample to 5000 cells
cs <- sample(colnames(so_merged), 5e3)
sub <- subset(so_merged, cells = cs)

## plot the expression of the endogenous TDP-43, TDP-HA, total TDP and the construct
tdp <- c("ENSG00000120948.TARDBP", "ENSG00000120948.TARDBP-alevin", "TDP43-HA", "total-TDP43", "construct")
names(tdp) <- c("TARDBP-CellRanger", "TARDBP-alevin", "TDP-HA", "total-TDP43", "construct")
cat("## TDP-43\n")
ps <- lapply(seq_along(tdp), function(i) {
    if (!tdp[i] %in% rownames(sub)) return(NULL)
    FeaturePlot(sub, features = tdp[i], reduction = "umap", pt.size = 0.4,
                slot = "data") +
        theme(aspect.ratio = 1, legend.position = "none") +
        ggtitle(names(tdp)[i]) + theme_void() + theme(aspect.ratio = 1)
})
# arrange plots in grid
ps <- ps[!vapply(ps, is.null, logical(1))]
p <- plot_grid(plotlist = ps, ncol = 4, label_size = 10)
print(p)
cat("\n\n")

## control genes STMN2 and VIM
g <- c("ENSG00000104435.STMN2", "STMN2-alevin", "ENSG00000026025.VIM", "VIM-alevin")
names(g) <- c("STMN2-CellRanger", "STMN2-alevin", "VIM-CellRanger", "VIM-alevin")
cat("## control genes alevin\n")
ps <- lapply(seq_along(g), function(i) {
    if (!g[i] %in% rownames(sub)) return(NULL)
    FeaturePlot(sub, features = g[i], reduction = "umap", pt.size = 0.4,
                slot = "data") +
        theme(aspect.ratio = 1, legend.position = "none") +
        ggtitle(names(g)[i]) + theme_void() + theme(aspect.ratio = 1)
})
# arrange plots in grid
ps <- ps[!vapply(ps, is.null, logical(1))]
p <- plot_grid(plotlist = ps, ncol = 4, label_size = 10)
print(p)
cat("\n\n")
```


# Heatmap with TDP, construct and control genes

```{r prepare-sce-merged}
## prepare sce object with all genes in one count matrix!
sce_merged <- as.SingleCellExperiment(so_merged)
sce_merged$cluster_id <- Idents(so_merged)
sce_merged$sample_id <- so_merged$sample_id
```

```{r known-markers-heatmap-merged, fig.width = 10, fig.height = 18}
# split cells by cluster
cs_by_k <- split(colnames(sce_merged), sce_merged$cluster_id)
# compute cluster-marker means
ms_by_cluster <- lapply(fs, function(gs) vapply(cs_by_k, function(i)
        Matrix::rowMeans(logcounts(sce_merged)[gs, i, drop = FALSE]), 
        numeric(length(gs))))

# prep. for plotting & scale b/w 0 and 1
mat <- do.call("rbind", ms_by_cluster)
mat <- muscat:::.scale(mat)
rownames(mat) <- gs
cols <- muscat:::.cluster_colors[seq_along(fs)]
cols <- setNames(cols, names(fs))
row_anno <- rowAnnotation(
    df = data.frame(label = factor(ks, levels = names(fs))),
    col = list(label = cols), gp = gpar(col = "white"))
# percentage of cells from each of the samples per cluster
(n_cells <- table(sce_merged$cluster_id, sce_merged$sample_id))
sample_props <- prop.table(n_cells, margin = 1)
col_mat <- as.matrix(unclass(sample_props))
sample_cols <- c("#882255", "#11588A",  "#117733", "#44AA99")
sample_cols <- setNames(sample_cols, colnames(col_mat))
col_anno <- HeatmapAnnotation(
    perc_sample = anno_barplot(col_mat, gp = gpar(fill = sample_cols), 
                               height = unit(2, "cm"),
                               border = FALSE),
    annotation_label = "fraction of sample\nin cluster",
    gap = unit(10, "points"))
col_lgd <- Legend(labels = names(sample_cols),
       title = "sample",
       legend_gp = gpar(fill = sample_cols))

hm <- Heatmap(mat,
    name = "scaled avg.\nexpression",
    col = viridis(10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    column_title = "cluster_id",
    column_title_side = "bottom",
    column_names_side = "bottom",
    column_names_rot = 0, 
    column_names_centered = TRUE,
    rect_gp = gpar(col = "white"),
    left_annotation = row_anno,
    top_annotation = col_anno)
draw(hm, annotation_legend_list = list(col_lgd))
```


# Dotplot

```{r dotplot, fig.width = 8, fig.height = 4}
features <- c("ENSG00000120948.TARDBP", "ENSG00000120948.TARDBP-alevin", "TDP43-HA", "total-TDP43", "construct", "ENSG00000104435.STMN2", "STMN2-alevin", "ENSG00000026025.VIM", "VIM-alevin")
fs <- c("TARDBP-CellRanger", "TARDBP-alevin", "TDP-HA", "total-TDP43", "construct", "STMN2-CellRanger", "STMN2-alevin", "VIM-CellRanger", "VIM-alevin")

DotPlot(so_merged, assay = "RNA", features = features, 
        scale = TRUE, scale.min = 0, scale.max = 100, dot.scale = 7) + 
  RotatedAxis() + scale_color_viridis() + 
  theme(axis.text.x = element_text(angle=45)) + ylab("cluster ID") + 
  scale_x_discrete(name = "gene", breaks = features, labels=fs)

## only the neuronal clusters
neuronal_clusters <- c(0, 2:5, 7:11, 12)
DotPlot(so_merged, assay = "RNA", features = features, idents = neuronal_clusters, 
        scale = TRUE, scale.min = 0, scale.max = 100, dot.scale = 7) + 
  RotatedAxis() + scale_color_viridis() + 
  theme(axis.text.x = element_text(angle=45)) + ylab("cluster ID") + 
  scale_x_discrete(name = "gene", breaks = features, labels=fs)
```


# Check for logFC of specific genes

We first subset the data to the neuronal clusters.
```{r subset-neuronal-clusters}
sce_sub <- sce_merged[,sce_merged$cluster_id %in% neuronal_clusters]
```

```{r scran-findMarkers, warning = FALSE}
## markers for all other clusters
all_default <- findMarkers(sce_sub, groups = sce_sub$cluster_id, 
                           pval.type="all", assay.type = "logcounts")
all_default[["12"]][1:10,]

## what is the logFC for the different TDP quantifications and our control genes?
all_default[["12"]][which(rownames(all_default[["12"]]) %in% features),]

## compute the mean logFC, because the reporte summary logFC is the smallest logFC to any cluster
rowMeans(all_default[["12"]][which(rownames(all_default[["12"]]) %in% features),] %>% 
           as.data.frame %>% dplyr::select(starts_with("logFC")), na.rm = TRUE)
```


```{r seurat-foldchange}
## using Seurat
FoldChange(object = so_merged, slot = "data", ident.1 = 12, 
            ident.2 = neuronal_clusters, 
            features = features, pseudocount.use = 1)
# the default pseudocount is 1, but that gives much higher logFC than scran findMarkers
# a value of 3 gives similar results...
FoldChange(object = so_merged, slot = "data", ident.1 = 12, 
            ident.2 = neuronal_clusters, 
            features = features, pseudocount.use = 3)
# a small pseudocount doesn't not affect genes with low expression that much
FoldChange(object = so_merged, slot = "data", ident.1 = 12, 
            ident.2 = neuronal_clusters, 
            features = features, pseudocount.use = 0.001)
```

## Save cluster markers to RDS

```{r save-rds}
saveRDS(so_merged, file.path("output", "so_TDP-06-01-totalTDP-construct-quantification.rds"))
saveRDS(sce_merged, file.path("output", "sce_TDP-06-01-totalTDP-construct-quantification.rds"))
```

