---
title: "Plasmid expression"
author: "Katharina Hembach"
date: "10/14/2020"
output: 
  html_document:
    toc: true,
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```


### Load packages

```{r, message = FALSE}
library(cowplot)
library(ggplot2)
library(Seurat)
library(SingleCellExperiment)
library(tximport)
```

## Load data

We combine the quantification of the plasmid transcript and the endogenous TDP-43 with the CellRanger count matrix.

```{r read-sce}
sce <- readRDS(file.path("output", "sce_TDP_03_filtering.rds"))
## we only keep the four samples of the TDP-43 experiment
sample_ids <- c("TDP4wOFF", "TDP2wON", "TDP4wONa", "TDP4wONb")
sce <- sce[,colData(sce)$sample_id %in% sample_ids]
sce$sample_id <- droplevels(sce$sample_id)
```

```{r read-alevin}
samples <- c("no1_Neural_cuture_d_96_TDP-43-HA_4w_DOXoff", 
             "no2_Neural_cuture_d_96_TDP-43-HA_2w_DOXON",
             "no3_Neural_cuture_d_96_TDP-43-HA_4w_DOXONa",
             "no4_Neural_cuture_d_96_TDP-43-HA_4w_DOXONb")
txi <- matrix(NA, nrow = 2)
for (i in 1:4) {
  fi <- file.path("data", "Sep2020", "alevin_TDP43", samples[i], 
                     "alevin/quants_mat.gz")

  # import alevin quants
  a <- tximport(fi, type="alevin")$counts
  
  ## match the alevin and CellRanger cell IDs
  colnames(a) <- paste0(colnames(a), "-1.", sample_ids[i])
  txi <- cbind(txi, a)
}
txi <- txi[,colnames(txi) != ""]
rownames(txi) <- c("ENSG00000120948.TARDBP-alevin", "TDP43-HA")
```

We add the alevin counts to the CellRanger matrix.
```{r combine-counts}
## add two new rows to counts matrix and replace the counts for matching 
## barcodes with the alevin counts
counts <- rbind(counts(sce), DelayedArray(matrix(0, nrow = 2, 
                                                 ncol = ncol(counts(sce)))))
rownames(counts) <- c(rownames(sce), rownames(txi))
## match the barcodes
m <- match(colnames(txi), colnames(sce))
counts[rownames(txi),m[!is.na(m)]] <- txi[,which(!is.na(m))]

so <- CreateSeuratObject(
    counts = counts,
    meta.data = data.frame(colData(sce)),
    project = "TDP_experiment")
```

## Normalization
```{r normalization, warning = FALSE}
# split by sample
cells_by_sample <- split(colnames(sce), sce$sample_id)
so <- lapply(cells_by_sample, function(i) subset(so, cells = i))

## log normalize the data using a scaling factor of 10000
so <- lapply(so, NormalizeData, verbose = FALSE, scale.factor = 10000, 
             normalization.method = "LogNormalize")
```

We merge the normalized and data of the six samples into a combined Seurat object and compute variable features.
```{r merge-samples}
so_list <- so
## merge the individial Seurat objects and conserve the normalized and scaled data
so <- merge(so[[1]], y = so[2:length(so)], project = "TDP_experiment", 
            merge.data = TRUE)
```

```{r variable-features-scaling}
so <- FindVariableFeatures(so, nfeatures = 2000, 
    selection.method = "vst", verbose = FALSE)
so <- ScaleData(so, verbose = FALSE, vars.to.regress = c("sum", 
                                                         "subsets_Mt_percent"))
```


## Dimension reduction 

We perform dimension reduction with t-SNE and UMAP based on PCA results.
```{r dimension-reduction, warning = FALSE}
so <- RunPCA(so, npcs = 30, verbose = FALSE)
so <- RunTSNE(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, do.fast = TRUE, verbose = FALSE)
so <- RunUMAP(so, reduction = "pca", dims = seq_len(20),
    seed.use = 1, verbose = FALSE)
```

### Plot PCA results

```{r, fig.width = 12, fig.height = 8}
# top genes that are associated with the first two PCs
VizDimLoadings(so, dims = 1:2, reduction = "pca")
```

```{r, fig.width = 10, fig.height = 8}
## PCA plot 
DimPlot(so, reduction = "pca", group.by = "sample_id")
```

```{r}
# elbow plot with the ranking of PCs based on the % of variance explained
ElbowPlot(so, ndims = 30)
```


## Clustering

We cluster the cells using the reduced PCA dimensions.

```{r clustering, warning = FALSE}
so <- FindNeighbors(so, reduction = "pca", dims = seq_len(20), verbose = FALSE)
for (res in c(0.2, 0.4, 0.8, 1))
    so <- FindClusters(so, resolution = res, random.seed = 1, verbose = FALSE)
```


## Dimension reduction plots 

We plot the dimension reduction (DR) and color by sample, group and cluster ID
```{r dr-plots, fig.width = 10, fig.height = 9, warning = FALSE}
thm <- theme(aspect.ratio = 1, legend.position = "none")
ps <- lapply(c("sample_id", "ident"), function(u) {
    p1 <- DimPlot(so, reduction = "tsne", group.by = u) + thm
    p2 <- DimPlot(so, reduction = "umap", group.by = u)
    lgd <- get_legend(p2)
    p2 <- p2 + thm
    list(p1, p2, lgd)
    plot_grid(p1, p2, lgd, nrow = 1,
        rel_widths = c(1, 1, 0.5))
})
plot_grid(plotlist = ps, ncol = 1)
```


## QC on DR plots {.tabset}

```{r DR-QC, results = "asis", fig.width = 12}
cs <- sample(colnames(so), 1e4) ## subsample cells
.plot_features <- function(so, dr, id) {
    FeaturePlot(so, cells = cs, features = id, reduction = dr, pt.size = 0.4, 
                cols = c("grey", "blue")) +
        guides(col = guide_colourbar()) +
        theme_void() + theme(aspect.ratio = 1)
}
ids <- c("sum", "detected", "subsets_Mt_percent", "ENSG00000120948.TARDBP", 
         "ENSG00000120948.TARDBP-alevin", "TDP43-HA")
for (id in ids) {
    cat("### ", id, "\n")
    p1 <- .plot_features(so, "tsne", id)
    lgd <- get_legend(p1)
    p1 <- p1 + theme(legend.position = "none") + ggtitle("tSNE")
    p2 <- .plot_features(so, "umap", id) + theme(legend.position = "none") + 
      ggtitle("UMAP")
    ps <- plot_grid(plotlist = list(p1, p2), nrow = 1)
    p <- plot_grid(ps, lgd, nrow = 1, rel_widths = c(1, 0.2))
    print(p)
    cat("\n\n")
}
```

## Save Seurat object to RDS

```{r}
saveRDS(so, file.path("output", "so_TDP_05_plasmid_expression.rds"))
```

