---
title: "Cluster analysis"
author: "Katharina Hembach"
date: "4/7/2021"
output: 
  html_document:
    toc: true,
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep = TRUE, cache = TRUE, dev = "png",
                      dev.args = list(png = list(type = "cairo")), 
                      message = FALSE, cache.lazy = FALSE)
```

### Load packages

```{r load-libs, message = FALSE, warning = FALSE}
library(ComplexHeatmap)
library(cowplot)
library(ggplot2)
library(dplyr)
library(muscat)
library(RColorBrewer)
library(Seurat)
library(SingleCellExperiment)
```

# Load data & convert to SCE

```{r load-data}
so <- readRDS(file.path("output", "so_timepoints-01-organoid_integration.rds"))
sce <- as.SingleCellExperiment(so, assay = "RNA")
colData(sce) <- as.data.frame(colData(sce)) %>% 
    mutate_if(is.character, as.factor) %>% 
    DataFrame(row.names = colnames(sce))
levels(sce$integration_group)[levels(sce$integration_group) == "NC"] <- "iCoMoNSC"
sce$integration_group <- relevel(sce$integration_group, "iCoMoNSC")
```

# Cluster-sample counts

```{r}
(n_cells <- table(sce$cluster_id, sce$integration_group))
```

# Relative cluster-abundances

```{r fig.width = 5, fig.height = 8}
fqs <- prop.table(n_cells, margin = 2)
mat <- round(as.matrix(unclass(fqs))*100, 2)
colfunc <- colorRampPalette(c("ghostwhite", "deepskyblue4"))
Heatmap(mat,
    col = colfunc(10),
    name = "Percentage\nof cells",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster ID",
    column_title = "cell line",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(mat[j, i] , x = x, y = y, 
            gp = gpar(col = "black", fontsize = 10)))
```

```{r fig.width = 7, fig.height = 7}
(n_cells_group <- table(sce$cluster_id, sce$Stage))
fqs <- prop.table(n_cells_group, margin = 2)
mat <- round(as.matrix(unclass(fqs))*100, 2)
Heatmap(mat,
    col = colfunc(10),
    name = "Percentage\nof cells",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cluster ID",
    column_title = "time point",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(mat[j, i] , x = x, y = y, 
            gp = gpar(col = "black", fontsize = 10)))
```

Heatmap with the lineage labels from the Kanton et al. paper. Cells without the label (all iCoMoNSCs) are excluded from the heatmap!

```{r fig.width = 8, fig.height = 8}
(n_cells_lineage <- table(sce$cluster_id, sce$cl_FullLineage))
fqs <- prop.table(n_cells_lineage, margin = 2)
mat <- round(as.matrix(unclass(fqs))*100, 2)
cn <- colnames(mat)
Heatmap(mat,
    col = colfunc(10),
    name = "Percentage\nof cells",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_column_names = FALSE, 
    row_names_side = "left",
    row_title = "cluster ID",
    column_title = "Lineage",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(mat[j, i], x = x, y = y, 
            gp = gpar(col = "black", fontsize = 10)),
    bottom_annotation = HeatmapAnnotation(
      text = anno_text(cn, rot = 80, just = "right")))

(n_cells_lineage <- table(sce$cl_FullLineage, sce$cluster_id))
fqs <- prop.table(n_cells_lineage, margin = 2)
mat <- round(as.matrix(unclass(fqs))*100, 2)
Heatmap(mat,
    col = colfunc(10),
    name = "Percentage\nof cells",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "Lineage",
    row_names_rot = 10,
    column_title = "cluster ID",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(mat[j, i], x = x, y = y, 
            gp = gpar(col = "black", fontsize = 10)))
```

Heatmap with the line labels from the Kanton et. al paper. Cell without the label, are excluded from the heatmap!

```{r fig.width = 8, fig.height = 8}
(n_cells_linecomp <- table(sce$cluster_id, sce$cl_LineComp))
fqs <- prop.table(n_cells_linecomp, margin = 2)
mat <- round(as.matrix(unclass(fqs))*100, 2)
cn <- colnames(mat)
Heatmap(mat,
    col = colfunc(10),
    name = "Percentage\nof cells",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_column_names = FALSE, 
    row_names_side = "left",
    row_title = "cluster ID",
    column_title = "line ",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(mat[j, i], x = x, y = y, 
            gp = gpar(col = "black", fontsize = 10)),
    bottom_annotation = HeatmapAnnotation(
      text = anno_text(cn, rot = 80, just = "right")))

(n_cells_linecomp <- table(sce$cl_LineComp, sce$cluster_id))
fqs <- prop.table(n_cells_linecomp, margin = 2)
mat <- round(as.matrix(unclass(fqs))*100, 2)
Heatmap(mat,
    col = colfunc(10),
    name = "Percentage\nof cells",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "Lineage",
    row_names_rot = 10,
    column_title = "line",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(mat[j, i], x = x, y = y, 
            gp = gpar(col = "black", fontsize = 10)))
```

Heatmap with the cell type labels from the Kanton et. al paper. Cell without the label, are excluded from the heatmap!

```{r fig.width = 8, fig.height = 8}
(n_cells_predcell <- table(sce$cluster_id, sce$PredCellType))
fqs <- prop.table(n_cells_predcell, margin = 2)
mat <- round(as.matrix(unclass(fqs))*100, 2)
cn <- colnames(mat)
Heatmap(mat,
    col = colfunc(10),
    name = "Percentage\nof cells",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_column_names = FALSE, 
    row_names_side = "left",
    row_title = "cluster ID",
    column_title = "cell type",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(mat[j, i], x = x, y = y, 
            gp = gpar(col = "black", fontsize = 10)),
    bottom_annotation = HeatmapAnnotation(
      text = anno_text(cn, rot = 80, just = "right")))

(n_cells_predcell <- table(sce$PredCellType, sce$cluster_id))
fqs <- prop.table(n_cells_predcell, margin = 2)
mat <- round(as.matrix(unclass(fqs))*100, 2)
Heatmap(mat,
    col = colfunc(10),
    name = "Percentage\nof cells",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_names_side = "left",
    row_title = "cell type",
    row_names_rot = 10,
    column_title = "line",
    column_title_side = "bottom",
    rect_gp = gpar(col = "white"),
    cell_fun = function(i, j, x, y, width, height, fill)
        grid.text(mat[j, i], x = x, y = y, 
            gp = gpar(col = "black", fontsize = 10)))
```

